### 2022-01-13 README.Rmd
initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# maualithicspaper

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/maualithicspaper/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-01-21 analysis/paper/paper.Rmd
Added my name

---
title: "Title Goes Here"
author:
  - Quinn Habedank:
      email: fl@oneoeurhg.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-01-21 analysis/paper/paper.Rmd
add my name

@@
       correspondence: true
-  - Yourfirstname Yourlastname:
+  - Ben Marwick:
       email: fl@another.edu

### 2022-01-25 analysis/paper/paper.Rmd
test

@@
 
+# Abstract
+
+Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly
+patterned forms, which can make them challenging to analyze and interpret. As a result,
+many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly
+understood. We use 2D shape data to hypothesize a relationship between unretouched
+flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeo-
+logical site in northern Vietnam. We apply a Principal Components Analysis to the flake
+outlines to investigate shape variation throughout the reduction sequence (measured
+by dorsal cortex coverage). We find that flake shape varies by reduction stage, primarily
+through differences in flake length and width. Our results suggest that flake shape is
+sensitive to assemblage reduction intensity, and may give useful comparative insights
+when other attributes show little variation. These results are important for understanding
+stone artefact assemblages from Southeast Asia which often yield little variation when
+analysed with traditional approaches.
+
 # Introduction
+
+- Flake shape is often related to the uses of said flakes (insert citations here)
+- Stone tools from South East Asia tend to lack the distinctive outlines of European tools (insert citations here)
+- Flake outline analysis was condu
 

### 2022-01-28 analysis/paper/paper.Rmd
Updated outline

@@
 # Introduction
-
-- Flake shape is often related to the uses of said flakes (insert citations here)
-- Stone tools from South East Asia tend to lack the distinctive outlines of European tools (insert citations here)
-- Flake outline analysis was condu
+- Stone tools from South East Asia tend to lack distinctive typological categories (Mijares 2008; Borel et al. 2013; Borel et al. 2017)
+- The purpose of this study was to examine the relationship between flake shape and reduction intensity
+- Approach used in the paper is a response to the lack of traditional typological categories in South East Asia
 
@@
 # Background
-
+- Mau A is an open air site located near the Red River in Yen Bai Province, Northern Vietnam.
+- Excavation conducted in 2015 as a part of a UW field school
+- A wide array of flakes and stone tools were recovered, n=1058
+- Data for the paper comes from caliper measurements of flakes from the 2015 excavation
 # Methods
-
+(Note, this is copied largely from the poster, and will require further editing -QH)
+-In order to gather the 1058 stone artefacts, excavations were conducted using hand tools and sieved 4 square meter plots of excavated sediment. 
+-A custom function was used to transform artefact measurements into outlines
+-PCA was conducted on the landmarks outlines
+-Reduction categories were based upon dorsal cortex percentage.
+Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995)
 # Results
@@
 # References 
+Borel, Antony, Richard Cornette, and Michel Baylac.
+2017.
+“Stone Tool Forms and Functions: A Morphometric Analysis of Modern Humans’ Stone Tools From Song Terus Cave (Java, Indonesia).” Archaeometry 59 (3): 455–71.
+https://doi.org/10.1111/arcm.12264.
 
+Borel, Antony, Claire Gaillard, Marie-Hélène Moncel, Robert Sala, Emmanuelle Pouydebat, Truman Simanjuntak, and François Sémah.
+2013.
+“How to Interpret Informal Flakes Assemblages?
+Integrating Morphological Description, Usewear and Morphometric Analysis Gave Better Understanding of the Behaviors of Anatomically Modern Human from Song Terus (Indonesia).” Journal of Anthropological Archaeology 32 (4): 630–46.
+https://doi.org/10.1016/j.jaa.2013.03.002.
+
+Bradbury, Andrew P., and Philip J.
+Carr.
+"Flake Typologies and Alternative Approaches: An Experimental Assessment." Lithic Technology 20, no.
+2 (1995): 100-15.
+Accessed June 11, 2021.
+http://www.jstor.org/stable/23273168.
+
+Mijares, A. 2008. “The late pleistocene to early holocene foragers of northern Luzon.”		Bulletin of the Indo-Pacific Prehistory Association 28: 99-107.
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

### 2022-01-28 analysis/paper/paper.Rmd
minor edits to add white space

@@
 
-# Introduction
+## Introduction
+
 - Stone tools from South East Asia tend to lack distinctive typological categories (Mijares 2008; Borel et al. 2013; Borel et al. 2017)
@@
 
-# Background
+## Background
+
 - Mau A is an open air site located near the Red River in Yen Bai Province, Northern Vietnam.
@@
 - Data for the paper comes from caliper measurements of flakes from the 2015 excavation
-# Methods
+
+## Methods
+
 (Note, this is copied largely from the poster, and will require further editing -QH)
@@
 Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995)
-# Results
+
+## Results
 
@@
 
-# References 
+## References
+
 Borel, Antony, Richard Cornette, and Michel Baylac.
@@
 Mijares, A. 2008. “The late pleistocene to early holocene foragers of northern Luzon.”		Bulletin of the Indo-Pacific Prehistory Association 28: 99-107.
+
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

### 2022-02-03 analysis/paper/paper.Rmd
methods update

--- 2022-01-28 analysis/paper/paper.Rmd
@@
 # Methods
+
+The data used for our analysis came from a 2015 dig at Mau A, located in Northern Vietnam, conducted by a University of Washington field school in conjunction with local archaeologists.
+Raw flake dimension measurements, such as max dimension length and width at various points along the max dimension length, were converted into shape outlines using an R script.
+
 (Note, this is copied largely from the poster, and will require further editing -QH)

### 2022-02-04 analysis/paper/paper.Rmd
add title, add numerous notes and comments, add text about excavation

--- 2022-01-28 analysis/paper/paper.Rmd
@@
 ---
-title: "Title Goes Here"
+title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Norther Vietnam"
 author:
@@
 
-Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly
-patterned forms, which can make them challenging to analyze and interpret. As a result,
-many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly
-understood. We use 2D shape data to hypothesize a relationship between unretouched
-flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeo-
-logical site in northern Vietnam. We apply a Principal Components Analysis to the flake
-outlines to investigate shape variation throughout the reduction sequence (measured
-by dorsal cortex coverage). We find that flake shape varies by reduction stage, primarily
-through differences in flake length and width. Our results suggest that flake shape is
-sensitive to assemblage reduction intensity, and may give useful comparative insights
-when other attributes show little variation. These results are important for understanding
-stone artefact assemblages from Southeast Asia which often yield little variation when
-analysed with traditional approaches.
+Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret.
+As a result, many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly understood.
+We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam.
+We apply a Principal Components Analysis to the flake outlines to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
+We find that flake shape varies by reduction stage, primarily through differences in flake length and width.
+Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation.
+These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with traditional approaches.
 
@@
 
-## Background
+<!-- BM: In this section we need to review previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->
 
-- Mau A is an open air site located near the Red River in Yen Bai Province, Northern Vietnam.
-- Excavation conducted in 2015 as a part of a UW field school
-- A wide array of flakes and stone tools were recovered, n=1058
-- Data for the paper comes from caliper measurements of flakes from the 2015 excavation
+## Excavations at Mau A
+
+Mau A is an open air archaeological site located on the banks above 5 m above the Red River in Yen Bai Province, Northern Vietnam, at the confluence of a small stream entering the river.
+Excavations were conducted in 2015 by a collaboration including researchers from the University of Washington, the Yen Bai Provincial Museum, the Institute of Archaeology in Hanoi, and the Vietnam National University - Social Sciences and Humanities University.
+An area of 2 x 2 m was excavated in 13 units of ten centimeters deep, through dense silty clay deposits to a depth of about 1.3 m below the surface.
+Subtle changes in stratigraphy indicated a slightly sloping deposit with four layers, consistent with reports of excavations in this location in the 1980s.
+A small amount of plastic and modern ceramics were found in the uppermost 0.1 m of deposit.
+The most striking feature identified during our excavations was the third layer, which is distinctive as very dense layer of flaked stone artefacts at about 0.6-1.1 m below the surface.
+No other features were identified during excavation.
+
+This dense lithic layer contained the majority of stone artefacts recovered from the site.
+The  chaîne opératoire of Mau A lithic technology is composed of unifacial shaping on long cobbles to produce sumatraloids (sumatralith-like pieces).
+The chaîne opératoire also involved the shaping of thick ovoid cobbles for the production of choppers or chopping-tools and half-cobbles (longitudinally split) that are shaped into tools with transverse cutting edges.
+Flake scar surfaces and edges are fresh and unweathered, indicating _in situ_ production and limited post-depositional movement.
+While previous work has claimed the Mau A assemblage is Son Vian (a precurser to Hoabinhian technology, characterised by unifacial flaking of blocky, cubic cobbles)
+
+<!-- BM: need to cite The Oxford Handbook of the Archaeology and Anthropology of Hunter-Gatherers, Ha Van Tan 1997, and the Proceedings of the Prehistoric Society -->
+
+Four radiocarbon ages were obtained from isolated charcoal fragments in the deposit.
+The charcoal was sampled to bracket the upper and lower boundaries of the dense lithic layer.
+<!-- add calibrated age numbers here --> The ages indicate a rapid accumulation process for this layer at the terminal Pleistocene.
+
+<!-- BM: add a paragraph on MH's pollen analysis to indicate local palaeoenvironment, and add her as co-author-->
 
@@
 
-(Note, this is copied largely from the poster, and will require further editing -QH)
--In order to gather the 1058 stone artefacts, excavations were conducted using hand tools and sieved 4 square meter plots of excavated sediment. 
--A custom function was used to transform artefact measurements into outlines
--PCA was conducted on the landmarks outlines
--Reduction categories were based upon dorsal cortex percentage.
-Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995)
+We initially recorded conventional linear dimension measurements, such as max dimension length and width at various points along the max dimension length, from all flakes recovered from the excavation.
+Then we converted these linear dimensions into landmarks points to represent the size and shape of each artefact.
+The specific linear measurements used to create the 2D landmarks were the top horizontal measurement, the max dimension length, width at a quarter of the maximum length, width at half of the maximum length, and width at three quarters of the maximum length.
+
+<!-- BM: Because our approach here is very novel, most readers will struggle to understand it.
+To help them, I think we will need a table that defines each of our dimensions, and a figure that shows an example of our landmark points on a typical artefact -->
+
+The outlines formed by the shape landmarks were then normalized for size, orientation, and rotation.
+We used Generalized Procrustes analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data.
+Our methodology for carrying out the PCA on the shape landmarks was inspired by Theska et al.
+(2020), which also performed 2D landmark data analysis using PCA.
+Finally, the shape landmark values were analysed by Principal Components Analysis.
+PCA biplots were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
+The first two primary components were then used to create biplots comparing the shape data to the flake reduction categories, excavation unit, and mass clusters, with the secondary variables being represented by colored outlines on the biplots.
+
+<!-- BM: Need to add citations to give the reader further information on GPA and shape analysis in archaeology.
+Need to add a sentence of definitino of what is GPA and what is PCA.
+Partly to guide the reader to more details, if they want it, and partly to demonstrate our expertise with the method -->
+
+Reduction categories were based upon dorsal cortex percentages were discretized based upon standard cut points.
+Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
+Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
+
+<!-- BM: Need to add citations and definition about univariate k-means -->
 
@@
 
-<!-- Here's some example analysis code: -->
+<!-- BM: basic stats on raw materials and change in technology by depth to satisfy readers about the basic details of the assemblage,
+
+<!-- BM: then QH's more complex analysis on flakes after the basic data -->
+
+<!-- BM: then previous more complex analysis on cores after QH's analysis of flakes -->
 
@@
 
+Theska, T., Sieriebriennikov, B., Wighard, S.
+S., Werner, M.
+S., & Sommer, R.
+J.
+(2020).
+Geometric morphometrics of microscopic animals as exemplified by model nematodes.
+Nature Protocols, 15(8), 2611–2644.
+https://doi.org/10.1038/s41596-020-0347-z
+
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

### 2022-02-04 README.Rmd
add continuous integration

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/maualithicspaper/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/maualithicspaper/master?urlpath=rstudio) [![.github/workflows/knit-in-docker.yml](https://github.com/benmarwick/maualithicspaper/actions/workflows/knit-in-docker.yml/badge.svg)](https://github.com/benmarwick/maualithicspaper/actions/workflows/knit-in-docker.yml)
 

### 2022-02-11 analysis/paper/paper.Rmd
x

@@
 
+<!--(10 Feb 2022) QH: Just putting all of the code and writing here. Not going to worry about formatting for now -->
+```{r}
+library(tidyverse)
+
+# import the data
+ma_flakes <- readxl::read_excel(here::here("data/UW Fieldschool Mau A 2015 post_excavation_season_April_2016.xlsx"),
+                                col_names = FALSE)
+
+# transpose so that columns are variables, rows are individual artefacts
+# and set col data types so we can compute on them
+ma_flakes_tr <-
+ma_flakes %>%
+  pivot_longer(-`...1`)  %>%
+  pivot_wider(names_from = `...1`,
+              values_from = value) %>%
+  mutate_all(parse_guess) %>%
+  # fix a few bad guesses
+  mutate(across(c(mass,
+                thickness_at_0.25_width,
+                thickness_at_0.75_width,
+                dorsal_cortex_perc,
+                dorsal_scar_initiation_count),
+         parse_number))
+
+# export this data frame to use with our analysis
+write_csv(ma_flakes_tr,
+          here::here("data/ma_flakes_tr.csv"))
+
+```
+
 ```{r get-data, eval = FALSE}

### 2022-02-11 analysis/paper/paper.Rmd
All of the code for dorsal cortext visualization is present. Debugging likily needed. Will write analysis soon.

@@
 <!--(10 Feb 2022) QH: Just putting all of the code and writing here. Not going to worry about formatting for now -->
+<!--(10 Feb 2022) QH: On a secondary note, I managed to cause an error with docker in record time. We will need to fix that later-->
+<!--(10 Feb 2022) QH: Lastly, I'm going to do my best to get the code working tonight. If that fails, I will still write up my interpretation using prior images of the visualization-->
+
 ```{r}
@@
           here::here("data/ma_flakes_tr.csv"))
+
+```
+```{r}
+# here we import the CSV file of caliper measurements, then compute on the
+# measurements to generate coordinates for a shape outline
+
+library(tidyverse)
+
+# import
+ma_flakes_tr <- read_csv(here::here("data/ma_flakes_tr.csv"))
+
+# custom function is from here, we have a local version in 999-functions.R
+# so we can work offline if we need to
+# https://gist.github.com/benmarwick/fac5754131c1fcc14d32b1658215f0e3
+
+source(here::here("code/999-functions.R"))
+
+polys <- polygon_shape_by_variables(data = ma_flakes_tr,       # our data frame
+                                    x = 'width_at_0.50_length',                # col for coord for x-axis of poly centers
+                                    y = 'max_dimension',       # col for coord for y-axis of poly centers
+
+                                    sf = 10,    # scale factor
+                                    top_horizontal_measurement =    'platform_width', # cols to use to compute poly verts
+                                    main_vertical_measurement =     'percussion_length',
+                                    upper_horizontal_measurement =  'width_at_0.25_length',
+                                    middle_horizontal_measurement = 'width_at_0.50_length',
+                                    bottom_horizontal_measurement = 'width_at_0.75_length')
+
+# take a look to ensure we are on the right track
+p <- ggplot() +
+  geom_polygon(data=polys,
+               aes(x=x,
+                   y=y,
+                   group = fill),
+               colour="black",
+               alpha = 0.5) +
+  coord_equal() +
+  theme_bw()
+p
+
+# make ready to convert to TPS format (thin plate spline)
+polys_lmk <-
+polys %>%
+  group_by(fill) %>%
+  mutate(landmark.number = row_number(fill),
+         specimen.ids  = fill) %>%
+  ungroup() %>%
+  select(-fill)
+
+# export as TPS file
+create_tps(polys_lmk,
+           here::here("data/landmarks.tps"))
+
+```
+```{r}
+ma_flakes_tr_with_groups <-
+  ma_flakes_tr %>%
+  mutate(dorsal_cortext_group = as.factor(
+           case_when(
+             between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
+             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
+             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
+           ))) %>% 
+ mutate(mass_group = as.factor(test_result_mass$cluster))
+# plot mass groups
+ggplot(ma_flakes_tr_with_groups)  +
+  aes(mass,
+      fill = mass_group) +
+  geom_histogram() +
+  scale_fill_viridis_d(name = "Mass\ncluster") +
+  xlab("Mass (g)") +
+  theme_minimal(base_size = 14) +
+  theme(panel.background = element_rect(fill='white',
+                                        colour = "white"),
+        plot.background = element_rect(fill='white',
+                                       colour = "white"))
 

### 2022-02-11 analysis/paper/paper.Rmd
Last edit for the night. Code is in place, still need to write up my analysis. Trying to weed out bugs.

@@
 ```
+
 ```{r}
@@
 
-# take a look to ensure we are on the right track
-p <- ggplot() +
-  geom_polygon(data=polys,
-               aes(x=x,
-                   y=y,
-                   group = fill),
-               colour="black",
-               alpha = 0.5) +
-  coord_equal() +
-  theme_bw()
-p
-
 # make ready to convert to TPS format (thin plate spline)
@@
 ```
+
+<!--(11 Feb 2022) QH: Currently it is 1:20 AM. I have been fiddling about with code and github for awhile. I plan on fixing everything when I wake up, and, failing that, Friday afternoon.-->
 ```{r}
@@
              between(dorsal_cortex_perc, 51, 100) ~ "Primary"
-           ))) %>% 
- mutate(mass_group = as.factor(test_result_mass$cluster))
+           ))) #%>% 
+ #QH here, not sure what this is doing for now, just putting it off to the side: mutate(mass_group = as.factor(test_result_mass$cluster))
 # plot mass groups

### 2022-02-11 analysis/paper/paper.Rmd
commit

@@
 
-<!--(10 Feb 2022) QH: Just putting all of the code and writing here. Not going to worry about formatting for now -->
-<!--(10 Feb 2022) QH: On a secondary note, I managed to cause an error with docker in record time. We will need to fix that later-->
-<!--(10 Feb 2022) QH: Lastly, I'm going to do my best to get the code working tonight. If that fails, I will still write up my interpretation using prior images of the visualization-->
+<!--QH: Just putting all of the code and writing here. Not going to worry about formatting for now -->
+<!--QH: On a secondary note, I managed to cause an error with docker in record time. We will need to fix that later-->
+<!--QH: Lastly, I'm going to do my best to get the code working tonight. If that fails, I will still write up my interpretation using prior images of the visualization-->
 
@@
 
-<!--(11 Feb 2022) QH: Currently it is 1:20 AM. I have been fiddling about with code and github for awhile. I plan on fixing everything when I wake up, and, failing that, Friday afternoon.-->
+<!--QH: I have been fiddling about with code and github for awhile. I plan on fixing everything when I wake up, and, failing that, Friday afternoon.-->
 ```{r}
@@
              between(dorsal_cortex_perc, 51, 100) ~ "Primary"
-           ))) #%>% 
- #QH here, not sure what this is doing for now, just putting it off to the side: mutate(mass_group = as.factor(test_result_mass$cluster))
+           ))) %>% 
+ mutate(mass_group = as.factor(test_result_mass$cluster))
 # plot mass groups

### 2022-02-11 code/999-functions.R

@@ -1,0 +1,179 @@
+
+#-----------------------------------------------------------------------------
+#' Polygon vertices for plotting computed from an object's linear dimensions
+#'
+#' This function takes a data frame (one row per object) of object dimensions,
+#' where each object is defined by five linear measurements, and computes coords
+#' of polygon vertices at given x and y values (taken from the data frame)
+#'
+#' @param data data frame with at least seven numeric columns
+#' @param x column for x position of polygon centers
+#' @param y column for y position of polygon centers
+#' @param j rows of data frame to subset, default is all
+#' @param sf scaling factor to change the size of the polygons, default in 10 (times smaller than actual values)
+#' @param top_horizontal_measurement value of linear measurement across the top of the polygon
+#' @param main_vertical_measurement value of linear measurement up-and-down the centre of the polygon
+#' @param upper_horizontal_measurement value of linear measurement horizontally across the polygon at 25 % of the main_vertical_measurement (starting at the top of main_vertical_measurement)
+#' @param middle_horizontal_measurement value of linear measurement horizontally across the polygon at 50 % of the main_vertical_measurement (starting at the top of main_vertical_measurement)
+#' @param bottom_horizontal_measurement value of linear measurement horizontally across the polygon at 75 % of the main_vertical_measurement (starting at the top of main_vertical_measurement)
+#'
+#' @return
+#' @export
+#'
+#' @examples
+#'
+#'
+#' n <- 10
+#' my_data <- data.frame(
+#'   xcord = rnorm(n), # col for coord for x-axis of poly centers
+#'   ycord = rnorm(n), # col for coord for y-axis of poly centers
+#'   sf = 10,
+#'   main_vertical_measurement = runif(n)      + 5,
+#'   top_horizontal_measurement = runif(n)     + 4,
+#'   upper_horizontal_measurement = runif(n)   + 4,
+#'   middle_horizontal_measurement = runif(n)  + 3,
+#'   bottom_horizontal_measurement = runif(n)  + 3
+#' )
+#'
+#' polys <- polygon_shape_by_variables(data = my_data,
+#'                                     x = 'xcord',
+#'                                     y = 'ycord',
[…]

### 2022-02-11 analysis/paper/paper.Rmd
add in code block that does the univariate k-means

@@
 <!--QH: I have been fiddling about with code and github for awhile. I plan on fixing everything when I wake up, and, failing that, Friday afternoon.-->
+
+```{r}
+library(Ckmeans.1d.dp)
+
+test_result_dorsal_cortex <- Ckmeans.1d.dp(ma_flakes_tr$dorsal_cortex_perc)
+
+plot(test_result_dorsal_cortex)
+
+# It seems that there is no clear clustering in the Dorsal Cortex values. I will have to use values from the literature to divide the cortex values
+
+# try mass:
+test_result_mass <- Ckmeans.1d.dp(ma_flakes_tr$mass)
+
+# plot(test_result_mass)
+# we got six clusters
+```
+
 ```{r}
@@
  mutate(mass_group = as.factor(test_result_mass$cluster))
+
 # plot mass groups

### 2022-02-11 analysis/paper/paper.Rmd
simplify file paths and remove duplicate data files

@@
 # import the data
-ma_flakes <- readxl::read_excel(here::here("data/UW Fieldschool Mau A 2015 post_excavation_season_April_2016.xlsx"),
+ma_flakes <- readxl::read_excel(here::here("analysis/data/raw_data/UW Fieldschool Mau A 2015 post_excavation_season_April_2016.xlsx"),
                                 col_names = FALSE)
@@
 write_csv(ma_flakes_tr,
-          here::here("data/ma_flakes_tr.csv"))
+          here::here("analysis/data/derived_data/ma_flakes_tr.csv"))
 
@@
 # import
-ma_flakes_tr <- read_csv(here::here("data/ma_flakes_tr.csv"))
+ma_flakes_tr <- read_csv(here::here("analysis/data/derived_data/ma_flakes_tr.csv"))
 
@@
 
-source(here::here("code/999-functions.R"))
+source(here::here("analysis/code/999-functions.R"))
 
@@
 create_tps(polys_lmk,
-           here::here("data/landmarks.tps"))
+           here::here("analysis/data/derived_data/landmarks.tps"))
 
@@
 
-plot(test_result_dorsal_cortex)
+# plot(test_result_dorsal_cortex)
 
@@
 ```
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
-
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
-```
-
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
-
-```{r demo-inline-code}
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
 

### 2022-02-17 analysis/paper/paper.Rmd
Added PCA code to create dorsal cortex biplots

@@
 
-<!--QH: Just putting all of the code and writing here. Not going to worry about formatting for now -->
-<!--QH: On a secondary note, I managed to cause an error with docker in record time. We will need to fix that later-->
-<!--QH: Lastly, I'm going to do my best to get the code working tonight. If that fails, I will still write up my interpretation using prior images of the visualization-->
+## PCA Analysis on Dorsal Cortex Groups
 
@@
 ```
+```{r}
 
-<!--QH: I have been fiddling about with code and github for awhile. I plan on fixing everything when I wake up, and, failing that, Friday afternoon.-->
+library(tidyverse)
+library(geomorph)
+library(Morpho)
+library(mclust)
+library(cluster)
+library(factoextra)
 
-```{r}
-library(Ckmeans.1d.dp)
+lands <- readland.tps(
+  here::here("data/derived_data/landmarks.tps"),
+  specID = "ID")
 
-test_result_dorsal_cortex <- Ckmeans.1d.dp(ma_flakes_tr$dorsal_cortex_perc)
+# drop outliers  identified in the previous script
+outlier_ids <- c(1152,  606, 1147, 1125,  756 ,1023,  457,
+                 813,  415,  877,  677, 1055,  239,  811,  601,  861,
+                  794,  103 , 598 ,1110 , 906 ,  59 ,   4)
 
-# plot(test_result_dorsal_cortex)
+lands_no_outliers <- lands[, , -outlier_ids]
+lands_no_outliers_names <- dimnames(lands_no_outliers)[[3]]
 
-# It seems that there is no clear clustering in the Dorsal Cortex values. I will have to use values from the literature to divide the cortex values
+sup.geo <- gpagen(lands_no_outliers,  ProcD=F, Proj = T)
 
-# try mass:
-test_result_mass <- Ckmeans.1d.dp(ma_flakes_tr$mass)
+# Principal component analysis (PCA) using a base R function
+pca.geo <- prcomp(two.d.array(sup.geo$coord))
 
-# plot(test_result_mass)
-# we got six clusters
-```
+# TODO: find a meaningful way to break this up into groups
+# pass these groups onto the PCA for drawing ellipses
 
-```{r}
+# this block of code below shows one way to discretize cortex percentage
+# into three groups. We need to explore other cutoffs, and other grouping
+# variables besides dorsal cortex. This code below is a good template for
+# starting these explorations
 ma_flakes_tr_with_groups <-
   ma_flakes_tr %>%
-  mutate(dorsal_cortext_group = as.factor(
+  slice(-outlier_ids) %>%
+  mutate(dorsal_cortext_group =
            case_when(
              between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
-             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
-             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
-           ))) %>% 
- mutate(mass_group = as.factor(test_result_mass$cluster))
+             between(dorsal_cortex_perc, 11, 60) ~ "Secondary",
+             between(dorsal_cortex_perc, 61, 100) ~ "Primary"
+           ))
 
-# plot mass groups
-ggplot(ma_flakes_tr_with_groups)  +
-  aes(mass,
-      fill = mass_group) +
-  geom_histogram() +
-  scale_fill_viridis_d(name = "Mass\ncluster") +
-  xlab("Mass (g)") +
-  theme_minimal(base_size = 14) +
-  theme(panel.background = element_rect(fill='white',
-                                        colour = "white"),
-        plot.background = element_rect(fill='white',
-                                       colour = "white"))
+# ggplot2-based plot of the PCA results using a factoextra function
+fviz_pca_ind(pca.geo,
+             geom.ind = "point", # show points without labels
+            #  line below sets the grouping column for colouring the points
+             habillage = ma_flakes_tr_with_groups$dorsal_cortext_group,
+             palette = "npg", #color palette, adjust to preferences
+             addEllipses = T, #adds ellipses or convex hulls
+             ellipse.type = "convex", #convex hulls
+             mean.point = F, #don’t show average point
+             legend.title = "SP") +  #adjust accordingly)
+             coord_fixed(ratio = 1) #adjustable for variation-weighted plots
 

### 2022-02-17 analysis/paper/paper.Rmd
Update to results description

@@
              mean.point = F, #don’t show average point
-             legend.title = "SP") +  #adjust accordingly)
+             legend.title = "Dorsal Cortex Class") +  #adjust accordingly)
              coord_fixed(ratio = 1) #adjustable for variation-weighted plots
@@
 ```
+<!-- QH: Later on, the language "the visualization above" will be changed to "Figure X." -->
 
+  The PCA biplot above shows flake shape (as defined within the first two primary dimensions) in relation to dorsal cortex class.
+The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
+Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slighly larger range of shape variation.
+Lastly, the heavily worked tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
+
+  It should also be noted that shape variation is not equal along both primary dimensions. 
 # Discussion

### 2022-02-17 analysis/paper/paper.Rmd
Update to results section description, second paragraph

@@
   It should also be noted that shape variation is not equal along both primary dimensions.
+For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first primary dimension.
+Slight differences are seen along the second primary dimension, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first primary dimension.
 # Discussion

### 2022-02-18 analysis/paper/paper.Rmd
Update to results, added code for primrary component composition and a few extra comments

@@
 <!-- QH: Later on, the language "the visualization above" will be changed to "Figure X." -->
+<!-- QH: Also, on a secondary note, I had to copy the data file into a different location. That means that there are redundant data files that will need to be cleaned up later -->
 
@@
 Slight differences are seen along the second primary dimension, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first primary dimension.
+
+<!-- QH: This code chunk below I copied from the old repo.
+The visualizations on dimension contributions are relevant to my second paragraph.
+I need to go back to some of my old work to see which values pertain to which measurement on the flakes, and to write up a paragraph or two elaborating on what I have already said.
+-->
+
+```{r}
+# let's look at all the PCs and how much each original # variable contributes to them
+
+#view how much variation is described by each PC (ordered from PC1 to last PC)
+enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
+
+#get meaningful PCs
+getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
+
+#estimate contribution (%) of landmarks on PC1
+PC1_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 1)
+# take a look at the plot
+PC1_contrib
+
+LM_contrib_PC1 <- data.frame(LM = rep(NA, nrow(PC1_contrib$data)/2),
+                             PC1_contrib = rep(NA, nrow(PC1_contrib$data)/2))
+for (i in 1:(nrow(PC1_contrib$data)/2)) {
+  LM_contrib_PC1$PC1_contrib[i] <- sum(PC1_contrib$data[(i*2-1):(i*2),]$contrib)
+  LM_contrib_PC1$LM[i] <- as.character(i)
+}
+
+# TODO: decode the x-axis labels on the scree plot so we can describe in
+# terms of actual flake dimension, write some sentences to describe and interpret
+
+# estimate contribution (%) of landmarks on PC2
+PC2_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 2)
+# take a look at the plot
+PC2_contrib
+
+LM_contrib_PC2 <- data.frame(LM = rep(NA, nrow(PC2_contrib$data)/2),
+                             PC2_contrib = rep(NA, nrow(PC2_contrib$data)/2))
+for (i in 1:(nrow(PC2_contrib$data)/2)) {
+  LM_contrib_PC2$PC2_contrib[i] <- sum(PC2_contrib$data[(i*2-1):(i*2),]$contrib)
+  LM_contrib_PC2$LM[i] <- as.character(i)
+}
+
+# examine landmarks with major contribution (%) to PC1 and PC2
+important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
+important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
+important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
+# here is a table that summarises the most important landmarks in the PCA
+important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
+
+```
+
 # Discussion

### 2022-02-18 analysis/paper/paper.Rmd
Added extra source for GPA in methods section, edit to methods section.

@@
 The outlines formed by the shape landmarks were then normalized for size, orientation, and rotation.
-We used Generalized Procrustes analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data.
+We used Generalized Procrustes Analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data.
+Generalized Procrustes Analysis (or GPA) is a
+
+Jung & Woo (2019)
+
 Our methodology for carrying out the PCA on the shape landmarks was inspired by Theska et al.
@@
 
+Jung, H., & Woo, E.
+J.
+(2019).
+Changes in mandibular ramus shape from the Neolithic to modern periods in Korea.
+International Journal of Osteoarchaeology, 29(4), 634–643.
+https://doi.org/10.1002/oa.2759
+
 Mijares, A. 2008. “The late pleistocene to early holocene foragers of northern Luzon.”		Bulletin of the Indo-Pacific Prehistory Association 28: 99-107.

### 2022-02-18 analysis/paper/paper.Rmd
remove duplicate data files and edit path

@@
 lands <- readland.tps(
-  here::here("data/derived_data/landmarks.tps"),
+  here::here("analysis/data/derived_data/landmarks.tps"),
   specID = "ID")
@@
 ```
+
 <!-- QH: Later on, the language "the visualization above" will be changed to "Figure X." -->

### 2022-02-18 analysis/paper/paper.Rmd
Update to methods

--- 2022-02-18 analysis/paper/paper.Rmd
@@
 We used Generalized Procrustes Analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data.
-Generalized Procrustes Analysis (or GPA) is a
+Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks.
+For an example of GPA being utilized to examine 2D landmarks in an archaeological context, refer to Jung & Woo (2019).
 
-Jung & Woo (2019)
-
+  Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis.
 Our methodology for carrying out the PCA on the shape landmarks was inspired by Theska et al.
 (2020), which also performed 2D landmark data analysis using PCA.
-Finally, the shape landmark values were analysed by Principal Components Analysis.
+PCA
+
   PCA biplots were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.

### 2022-02-18 analysis/paper/paper.Rmd
Added new source for methods regarding univariate k-means

@@
 (2020), which also performed 2D landmark data analysis using PCA.
-PCA
-
-  PCA biplots were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
-The first two primary components were then used to create biplots comparing the shape data to the flake reduction categories, excavation unit, and mass clusters, with the secondary variables being represented by colored outlines on the biplots.
+Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set.
+PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
+The first two primary components formulated the individual points on biplot, while the secondary variables (flake reduction categories, excavation unit, and mass clusters) were represented by colored outlines on the biplots
 
@@
 
+<!-- QH: Added more information on GPA and PCA, as well as an addition citation -->
+
   Reduction categories were based upon dorsal cortex percentages were discretized based upon standard cut points.
@@
 Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
+
+Papageorgiou (2020)
 
@@
 
+Papageorgiou, I. (2020). Ceramic investigation: How to perform statistical analyses. Archaeological and Anthropological Sciences, 12(9), 210. https://doi.org/10.1007/s12520-020-01142-x
+
 Theska, T., Sieriebriennikov, B., Wighard, S.

### 2022-02-18 analysis/paper/paper.Rmd
Updated methods to include more information regarding univariate k-means, as well as a citation

@@
 Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
-
-Papageorgiou (2020)
+Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
+For more information on univariate k-means, Papageorgiou (2020) examines the various different statistical techniques (including k-means) that can be applied in an archaeological setting, specifically in the study of ceramics.
 
 <!-- BM: Need to add citations and definition about univariate k-means -->
+
+<!-- QH: added more info on univariate k-means, as well as a citation  -->
 

### 2022-02-24 analysis/paper/paper.Rmd
Bug fix for NA in Biplot legend

@@
              between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
-             between(dorsal_cortex_perc, 11, 60) ~ "Secondary",
+             between(dorsal_cortex_perc, 1, 60) ~ "Secondary",
              between(dorsal_cortex_perc, 61, 100) ~ "Primary"

### 2022-02-24 analysis/paper/paper.Rmd
Further bigfixing with NA values in dorsal cortex PCA biplot, see comments

@@
 # starting these explorations
+
+# Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the intial set of data with NA values being passed through PCA. In the process of fixing
+
 ma_flakes_tr_with_groups <-
   ma_flakes_tr %>%
+  filter(!is.na(dorsal_cortex_perc)) %>% 
   slice(-outlier_ids) %>%
@@
              legend.title = "Dorsal Cortex Class") +  #adjust accordingly)
-             coord_fixed(ratio = 1) #adjustable for variation-weighted plots
+             coord_fixed(ratio = 1)+ #adjustable for variation-weighted plots
+  theme(plot.title = element_blank())
 

### 2022-02-24 analysis/paper/paper.Rmd
Ran into issue of rows with NA values for dorsal cortex perc being passed into the intial PCA calc. Uploading what I have now before proceeding further.
Changed "methodology" to just "method"

@@
   Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis.
-Our methodology for carrying out the PCA on the shape landmarks was inspired by Theska et al.
+Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al.
 (2020), which also performed 2D landmark data analysis using PCA.
@@
 
-# Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the intial set of data with NA values being passed through PCA. In the process of fixing
+# Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the initial set of data with NA values being passed through PCA. In the process of fixing
 

### 2022-02-24 analysis/paper/paper.Rmd
Changed "Reduction categories were based upon dorsal cortex percentages were discretized based upon standard cut points" to "Dorsal cortex percentages discretized upon standard cut points were used to create reduction categories" in methods section
Further edits to wording. Removed "It should also be noted that" in results section

@@
 
-  Reduction categories were based upon dorsal cortex percentages were discretized based upon standard cut points.
+Dorsal cortex percentages discretized upon standard cut points were used to create reduction categories.
 Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
@@
 
-  It should also be noted that shape variation is not equal along both primary dimensions.
+  Shape variation is not equal along both primary dimensions.
 For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first primary dimension.

### 2022-02-24 analysis/paper/paper.Rmd
Further changes to intial wording in methods section
Added bookdown references in text. Temporarily turned a code change filtering out NA values in the results section into an in-line comment so the code will run for now, albeit with "NA" still in the legend of the biplot

@@
 # starting these explorations
+```
 
+```{r cortex-biplot}
 # Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the initial set of data with NA values being passed through PCA. In the process of fixing
@@
   ma_flakes_tr %>%
-  filter(!is.na(dorsal_cortex_perc)) %>% 
+ # filter(!is.na(dorsal_cortex_perc)) %>% 
   slice(-outlier_ids) %>%
@@
 
-  The PCA biplot above shows flake shape (as defined within the first two primary dimensions) in relation to dorsal cortex class.
+<!--“biplot above” replace this figure \@ref(fig:code-block-name)  code-block-name is like this ```{r code-block-name} be sure to give them descriptive, simple names -->
+
+  Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two primary dimensions) in relation to dorsal cortex class.
 The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
 Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slighly larger range of shape variation.
-Lastly, the heavily worked tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
+Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
 

### 2022-02-24 analysis/paper/paper.Rmd
Replaced "primary dimension" with "principal component" throughout

@@
 
-  Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two primary dimensions) in relation to dorsal cortex class.
+  Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
 The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
@@
 
-  Shape variation is not equal along both primary dimensions.
-For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first primary dimension.
-Slight differences are seen along the second primary dimension, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first primary dimension.
+  Shape variation is not equal along both principal components.
+For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first principal component.
+Slight differences are seen along the second principal component, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first principal component.
 

### 2022-02-24 analysis/paper/paper.Rmd
Removed extra text outputs from PCA contrib plots in results section. Created plot grid using cowplot

@@
 #view how much variation is described by each PC (ordered from PC1 to last PC)
-enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
+# enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
 
 #get meaningful PCs
-getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
+# getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
 
@@
 # take a look at the plot
-PC1_contrib
+# PC1_contrib
 
@@
 # take a look at the plot
-PC2_contrib
+# PC2_contrib
 
@@
 # examine landmarks with major contribution (%) to PC1 and PC2
-important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
-important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
-important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
+# important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
+# important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
+# important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
 # here is a table that summarises the most important landmarks in the PCA
-important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
+# important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
+
+library(cowplot)
+
+plot_grid(PC1_contrib, 
+          PC2_contrib)
 

### 2022-02-24 analysis/paper/paper.Rmd
Fixed error in code with incorrect values for defining dorsal cortex perc categories

@@
              between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
-             between(dorsal_cortex_perc, 1, 60) ~ "Secondary",
-             between(dorsal_cortex_perc, 61, 100) ~ "Primary"
+             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
+             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
            ))

### 2022-02-24 analysis/paper/paper.Rmd
Added the code for the other two biplots (mass group and excavation unit) and fixed an error where outliers were removed from the shape data but not from the data frames used for biplot groups.

@@
 
-# import
+# import the data 
 ma_flakes_tr <- read_csv(here::here("analysis/data/derived_data/ma_flakes_tr.csv"))
@@
 
+# Adding extra groups needed for biplots, as well as removing the outliers brought up earlier
+
 ma_flakes_tr_with_groups <-
   ma_flakes_tr %>%
+  slice(-c(1152,  606, 1147, 1125,  756 ,1023,  457,
+                 813,  415,  877,  677, 1055,  239,  811,  601,  861,
+                  794,  103 , 598 ,1110 , 906 ,  59 ,   4)) %>% 
  # filter(!is.na(dorsal_cortex_perc)) %>% 
-  slice(-outlier_ids) %>%
   mutate(dorsal_cortext_group =
@@
            ))
+
+library(Ckmeans.1d.dp)
+
+test_result_mass <- Ckmeans.1d.dp(ma_flakes_tr_with_groups$mass)
+
+ma_flakes_tr_with_groups <-
+  ma_flakes_tr_with_groups %>% 
+  mutate(mass_group = as.factor(test_result_mass$cluster))
 
@@
 Slight differences are seen along the second principal component, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first principal component.
+
+```{r}
+# Excavation unit PCA
+# ggplot2-based plot of the PCA results using a factoextra function
+fviz_pca_ind(pca.geo,
+             geom.ind = "point", # show points without labels
+            #  line below sets the grouping column for colouring the points
+             habillage = as.factor(ma_flakes_tr_with_groups$Unit),
+             palette = viridis::cividis(13), # color palette, adjust to preferences
+             addEllipses = T, #adds ellipses or convex hulls
+             ellipse.type = "convex", #convex hulls
+             mean.point = F, #don’t show average point
+             legend.title = "Excavation\nunit",
+             title = "Change over time in flake shape") +  #adjust accordingly)
+             coord_fixed(ratio = 1)  + #adjustable for variation-weighted plots
+             theme(panel.background = element_rect(fill='white',
+                                        colour = "white"),
+              plot.background = element_rect(fill='white',
+                                       colour = "white")) 
+```
+
+<!-- Need to add in code that actually creates the mass groups -->
+
+```{r}
+
+# Mass PCA
+# ggplot2-based plot of the PCA results using a factoextra function
+fviz_pca_ind(pca.geo,
+             geom.ind = "point", # show points without labels
+            #  line below sets the grouping column for colouring the points
+             habillage = ma_flakes_tr_with_groups$mass_group,
+             palette = viridis::viridis(6), #color palette, adjust to preferences
+             addEllipses = T, #adds ellipses or convex hulls
+             ellipse.type = "convex", #convex hulls
+             mean.point = F, #don’t show average point
+             legend.title = "Mass\ncluster",  #adjust accordingly
+            title = "Flake mass and flake shape") + 
+             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
+             theme(panel.background = element_rect(fill='white',
+                                        colour = "white"),
+              plot.background = element_rect(fill='white',
+                                       colour = "white"))
+
+```
 

### 2022-02-25 analysis/paper/paper.Rmd
Reordered paragraphs in the methods section
Revised and added new citations for univariate k-means

@@
 
+  Dorsal cortex percentages discretized upon standard cut points were used to create reduction categories.
+Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
+Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
+
+  Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
+We used the Ckmeans.1d.dp package in R to perform our univariate k-means analysis (Wang & Song, 2011).
+The Ckmeans.1d.dp has been used for univariate clustering in applications outside of archaeology prior to our work (Song & Zhong, 2020).
+
   We initially recorded conventional linear dimension measurements, such as max dimension length and width at various points along the max dimension length, from all flakes recovered from the excavation.
@@
 <!-- QH: Added more information on GPA and PCA, as well as an addition citation -->
-
-Dorsal cortex percentages discretized upon standard cut points were used to create reduction categories.
-Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
-Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
-Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
-For more information on univariate k-means, Papageorgiou (2020) examines the various different statistical techniques (including k-means) that can be applied in an archaeological setting, specifically in the study of ceramics.
 
@@
 
-Papageorgiou, I. (2020). Ceramic investigation: How to perform statistical analyses. Archaeological and Anthropological Sciences, 12(9), 210. https://doi.org/10.1007/s12520-020-01142-x
+Song, M., & Zhong, H.
+(2020).
+Efficient weighted univariate clustering maps outstanding dysregulated genomic zones in human cancers.
+Bioinformatics, 36(20), 5027–5036.
+https://doi.org/10.1093/bioinformatics/btaa613
 
@@
 https://doi.org/10.1038/s41596-020-0347-z
+
+Wang, H., & Song, M. (2011). Ckmeans.1d.dp: Optimal k-means Clustering in One Dimension by Dynamic Programming. The R Journal, 3(2), 29–33.
 

### 2022-02-25 analysis/paper/paper.Rmd
Reordering of methods section, along with additional citations

@@
 Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks.
-For an example of GPA being utilized to examine 2D landmarks in an archaeological context, refer to Jung & Woo (2019).
+An example of GPA being applied to stone tool landmarks is Okumura & Araujo (2014).
 
   Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis.
+Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set.
+PCA has been used to investigate 2D (Riede & Pedersen, 2018; Hoggard et al., 2019; Radinović & Kajtez, 2021) and 3D (Archer et al., 2021) outlines of stone artefacts in the past.
 Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al.
 (2020), which also performed 2D landmark data analysis using PCA.
-Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set.
 PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
@@
 ## References 
+Archer, W., Djakovic, I., Brenet, M., Bourguignon, L., Presnyakova, D., Schlager, S., Soressi, M., & McPherron, S.
+P.
+(2021).
+Quantifying differences in hominin flaking technologies with 3D shape analysis.
+Journal of Human Evolution, 150, 102912.
+https://doi.org/10.1016/j.jhevol.2020.102912
 
@@
 
-Jung, H., & Woo, E.
-J.
+Hoggard, C.
+S., McNabb, J., & Cole, J.
+N.
 (2019).
-Changes in mandibular ramus shape from the Neolithic to modern periods in Korea.
-International Journal of Osteoarchaeology, 29(4), 634–643.
-https://doi.org/10.1002/oa.2759
+The Application of Elliptic Fourier Analysis in Understanding Biface Shape and Symmetry Through the British Acheulean.
+Journal of Paleolithic Archaeology, 2(2), 115–133.
+https://doi.org/10.1007/s41982-019-00024-6
 
 Mijares, A. 2008. “The late pleistocene to early holocene foragers of northern Luzon.”		Bulletin of the Indo-Pacific Prehistory Association 28: 99-107.
+
+Okumura, M., & Araujo, A.
+G.
+M.
+(2014).
+Long-term cultural stability in hunter–gatherers: A case study using traditional and geometric morphometric analysis of lithic stemmed bifacial points from Southern Brazil.
+Journal of Archaeological Science, 45, 59–71.
+https://doi.org/10.1016/j.jas.2014.02.009
+
+Radinović, M., & Kajtez, I.
+(2021).
+Outlining the knapping techniques: Assessment of the shape and regularity of prismatic blades using elliptic Fourier analysis.
+Journal of Archaeological Science: Reports, 38, 103079.
+https://doi.org/10.1016/j.jasrep.2021.103079
+
+Riede, F., & Pedersen, J. B. (2018). Late Glacial Human Dispersals in Northern Europe and Disequilibrium Dynamics. Human Ecology, 46(5), 621–632. https://doi.org/10.1007/s10745-017-9964-8
 

### 2022-02-25 analysis/paper/paper.Rmd
Update to results section, was unable to change the labels on the bar charts to human-meaningful values; added a small table with said human-meaningful values just for now

@@
 <!-- QH: Added more information on GPA and PCA, as well as an addition citation -->
-
-<!-- BM: Need to add citations and definition about univariate k-means -->
-
-<!-- QH: added more info on univariate k-means, as well as a citation  -->
 
@@
 
-```{r}
+```{r contribution_plots}
 # let's look at all the PCs and how much each original # variable contributes to them
@@
 
+# This is a temporary fix just to get the labels looking nice for now; there is most certainly a cleaner way to do this
+# My fix did not work. I will have to create a temp table showing the values for now
+
+# PC1_contrib[["plot_env"]][["xx"]] <- as.factor(c("X Value: Top Left",
+#                                              "Y Value: Top Left",
+#                                               "Y Value: Mid-Upper Left",
+#                                               "X Value: Mid-Lower Left",
+#                                               "Y Value: Mid-Lower Left",
+#                                               "X Value: Bottom Left",
+#                                               "Y Value: Bottom Left",
+#                                               "X Value: Distal Tip",
+#                                               "Y Value: Distal Tip",
+#                                               "X Value: Bottom Right",
+#                                               "Y Value: Bottom Right",
+#                                               "X Value: Mid-Lower Right",
+#                                               "Y Value: Mid-Lower Right",
+#                                               "X Value: Mid-Upper Right",
+#                                               "Y Value: Mid-Upper Right",
+#                                               "X Value: Top Right",
+#                                               "Y Value: Top Right"))
+#
+
+#iris$Species[iris$Species == 'virginica'] <- 'setosa'
+
+#PC1_contrib[["plot_env"]][["data"]][["name"]][PC1_contrib[["plot_env"]][["data"]][["name"]] == '5.X'] <- "X Value: Distal Tip"
+
+#PC1_contrib[["plot_env"]][["data_sum"]][["name"]] <- as.factor(c(1:18))
+
+output_meanings <-tibble(data.frame(c("X Value: Top Left",
+         "Y Value: Top Left",
+         "X Value: Mid-Upper Left",
+         "Y Value: Mid-Upper Left",
+         "X Value: Mid-Lower Left",
+         "Y Value: Mid-Lower Left",
+         "X Value: Bottom Left",
+         "Y Value: Bottom Left",
+         "X Value: Distal Tip",
+         "Y Value: Distal Tip",
+         "X Value: Bottom Right",
+         "Y Value: Bottom Right",
+         "X Value: Mid-Lower Right",
+         "Y Value: Mid-Lower Right",
+         "X Value: Mid-Upper Right",
+         "Y Value: Mid-Upper Right",
+         "X Value: Top Right",
+         "Y Value: Top Right"),
+       c("1X",
+         "1Y",
+         "2X",
+         "2Y",
+         "3X",
+         "3Y",
+         "4X",
+         "4Y",
+         "5X",
+         "5Y",
+         "6X",
+         "6Y",
+         "7X",
+         "7Y",
+         "8X",
+         "8Y",
+         "9X",
+         "9Y")))
+
+output_meanings <- setNames(output_meanings, c("Landmark Value", "Computer Output"))
+
 plot_grid(PC1_contrib, 
           PC2_contrib)
+ output_meanings
 
 ```
+
+  <!-- Could not figure out how to change the labels on the x axis for the two bar charts, so I included a small table with the human-meaningful values for now-->
 

### 2022-02-25 analysis/paper/paper.Rmd
Added explainations for the other two biplots, as well as minor edits to the plots themselves (removing the pre-generated titles and such)

@@
 
-```{r}
+```{r excavation-unit-biplot}
 # Excavation unit PCA
@@
               plot.background = element_rect(fill='white',
-                                       colour = "white")) 
+                                       colour = "white")) +
+    theme(plot.title = element_blank())
 ```
+<!-- This portion is lifted largely from the ISKM presentation. Might well reword it later, but it will do for now -->
 
-<!-- Need to add in code that actually creates the mass groups -->
+  Figure \@ref(fig:excavation-unit-biplot) shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from.
+Excavation unit one is near the surface, and unit 13 is near the base.
+The general picture here is that shape variation changes little through the deposit.
+Units with fewer flakes, such as 13, show lower shape variation.
 
-```{r}
+```{r mass-biplot}
 
@@
               plot.background = element_rect(fill='white',
-                                       colour = "white"))
+                                       colour = "white")) +
+    theme(plot.title = element_blank())
 
@@
 
-<!-- QH: This code chunk below I copied from the old repo.
-The visualizations on dimension contributions are relevant to my second paragraph.
-I need to go back to some of my old work to see which values pertain to which measurement on the flakes, and to write up a paragraph or two elaborating on what I have already said.
--->
+  In figure \@ref(fig:mass-biplot) we see that greater shape variation is found in the lightest mass classes.
+The two lightest mass clusters, one and two, are in purple and blue respectively.
+These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.
 

### 2022-02-25 analysis/paper/paper.Rmd
Drafted first half of discussion section, will need more work later

@@
 
-  <!-- Could not figure out how to change the labels on the x axis for the two bar charts, so I included a small table with the human-meaningful values for now-->
+  <!-- Could not figure out how to change the labels on the x axis for the two bar charts, so I included a small table with the human-meaningful values for now. This must be fixed later-->
+
+  Figure \@ref(fig:contribution_plots) shows the relative contribution of each of the original 18 variables put through our PCA.
+Looking at the supplementary table, we can see that the first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points.
+The second primary dimension is composed largely of the y-values for bottom-most and upper-most points.
 
 # Discussion
+  <!-- This portion partially consists of material from the ISKM presentation.
+Might want to reword things even more.
+Moreover, this section needs citations, which will bolster the overall content of this section.
+-->
+
+  Our data visualizations reveal some noteworthy trends regarding the degree of core reduction.
+Figure \@ref(fig:excavation-unit-biplot) shows very little variation over time, with lots of overlap between groups in the biplot.
+Units with fewer flakes, such as 13, show lower shape variation.
+This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time.
+By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:mass-biplot), where there is a high degree of shape variation in the lighest flakes.
+This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter.
+This is further supported by Figure \@ref(fig:cortex-biplot), where tertiary flakes with little dorsal cortex show the highest shape variation.
+This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.
+
+  Looking at \@ref(fig:contribution_plots),
 
@@
 ## References
+
 Archer, W., Djakovic, I., Brenet, M., Bourguignon, L., Presnyakova, D., Schlager, S., Soressi, M., & McPherron, S.

### 2022-02-25 analysis/paper/paper.Rmd
stash

--- 2022-02-25 analysis/paper/paper.Rmd
@@
 
-  Dorsal cortex percentages discretized upon standard cut points were used to create reduction categories.
-Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
-Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
-
-  Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
-We used the Ckmeans.1d.dp package in R to perform our univariate k-means analysis (Wang & Song, 2011).
-The Ckmeans.1d.dp has been used for univariate clustering in applications outside of archaeology prior to our work (Song & Zhong, 2020).
-
   We initially recorded conventional linear dimension measurements, such as max dimension length and width at various points along the max dimension length, from all flakes recovered from the excavation.
@@
 Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks.
-An example of GPA being applied to stone tool landmarks is Okumura & Araujo (2014).
+For an example of GPA being utilized to examine 2D landmarks in an archaeological context, refer to Jung & Woo (2019).
 
   Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis.
+Our methodology for carrying out the PCA on the shape landmarks was inspired by Theska et al.
+(2020), which also performed 2D landmark data analysis using PCA.
 Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set.
-PCA has been used to investigate 2D (Riede & Pedersen, 2018; Hoggard et al., 2019; Radinović & Kajtez, 2021) and 3D (Archer et al., 2021) outlines of stone artefacts in the past.
-Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al.
-(2020), which also performed 2D landmark data analysis using PCA.
 PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
@@
 <!-- QH: Added more information on GPA and PCA, as well as an addition citation -->
+
+  Reduction categories were based upon dorsal cortex percentages were discretized based upon standard cut points.
+Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
+Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
+Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
+For more information on univariate k-means, Papageorgiou (2020) examines the various different statistical techniques (including k-means) that can be applied in an archaeological setting, specifically in the study of ceramics.
+
+<!-- BM: Need to add citations and definition about univariate k-means -->
+
+<!-- QH: added more info on univariate k-means, as well as a citation  -->
 
@@
 
-# import the data 
+# import
 ma_flakes_tr <- read_csv(here::here("analysis/data/derived_data/ma_flakes_tr.csv"))
@@
 # starting these explorations
-```
-
-```{r cortex-biplot}
-# Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the initial set of data with NA values being passed through PCA. In the process of fixing
-
-# Adding extra groups needed for biplots, as well as removing the outliers brought up earlier
-
 ma_flakes_tr_with_groups <-
   ma_flakes_tr %>%
-  slice(-c(1152,  606, 1147, 1125,  756 ,1023,  457,
-                 813,  415,  877,  677, 1055,  239,  811,  601,  861,
-                  794,  103 , 598 ,1110 , 906 ,  59 ,   4)) %>% 
- # filter(!is.na(dorsal_cortex_perc)) %>% 
+  slice(-outlier_ids) %>%
   mutate(dorsal_cortext_group =
@@
              between(dorsal_cortex_perc, 0, 0) ~ "Tertiary",
-             between(dorsal_cortex_perc, 1, 50) ~ "Secondary",
-             between(dorsal_cortex_perc, 51, 100) ~ "Primary"
+             between(dorsal_cortex_perc, 11, 60) ~ "Secondary",
+             between(dorsal_cortex_perc, 61, 100) ~ "Primary"
            ))
-
-library(Ckmeans.1d.dp)
-
-test_result_mass <- Ckmeans.1d.dp(ma_flakes_tr_with_groups$mass)
-
-ma_flakes_tr_with_groups <-
-  ma_flakes_tr_with_groups %>% 
-  mutate(mass_group = as.factor(test_result_mass$cluster))
 
@@
              legend.title = "Dorsal Cortex Class") +  #adjust accordingly)
-             coord_fixed(ratio = 1)+ #adjustable for variation-weighted plots
-  theme(plot.title = element_blank())
+             coord_fixed(ratio = 1) #adjustable for variation-weighted plots
 
@@
 
-<!--“biplot above” replace this figure \@ref(fig:code-block-name)  code-block-name is like this ```{r code-block-name} be sure to give them descriptive, simple names -->
-
-  Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
+  The PCA biplot above shows flake shape (as defined within the first two primary dimensions) in relation to dorsal cortex class.
 The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
 Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slighly larger range of shape variation.
-Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
+Lastly, the heavily worked tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
 
-  Shape variation is not equal along both principal components.
-For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first principal component.
-Slight differences are seen along the second principal component, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first principal component.
-
-```{r}
-# Excavation unit PCA
-# ggplot2-based plot of the PCA results using a factoextra function
-fviz_pca_ind(pca.geo,
-             geom.ind = "point", # show points without labels
-            #  line below sets the grouping column for colouring the points
-             habillage = as.factor(ma_flakes_tr_with_groups$Unit),
-             palette = viridis::cividis(13), # color palette, adjust to preferences
-             addEllipses = T, #adds ellipses or convex hulls
-             ellipse.type = "convex", #convex hulls
-             mean.point = F, #don’t show average point
-             legend.title = "Excavation\nunit",
-             title = "Change over time in flake shape") +  #adjust accordingly)
-             coord_fixed(ratio = 1)  + #adjustable for variation-weighted plots
-             theme(panel.background = element_rect(fill='white',
-                                        colour = "white"),
-              plot.background = element_rect(fill='white',
-                                       colour = "white")) 
-```
-
-<!-- Need to add in code that actually creates the mass groups -->
-
-```{r}
-
-# Mass PCA
-# ggplot2-based plot of the PCA results using a factoextra function
-fviz_pca_ind(pca.geo,
-             geom.ind = "point", # show points without labels
-            #  line below sets the grouping column for colouring the points
-             habillage = ma_flakes_tr_with_groups$mass_group,
-             palette = viridis::viridis(6), #color palette, adjust to preferences
-             addEllipses = T, #adds ellipses or convex hulls
-             ellipse.type = "convex", #convex hulls
-             mean.point = F, #don’t show average point
-             legend.title = "Mass\ncluster",  #adjust accordingly
-            title = "Flake mass and flake shape") + 
-             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
-             theme(panel.background = element_rect(fill='white',
-                                        colour = "white"),
-              plot.background = element_rect(fill='white',
-                                       colour = "white"))
-
-```
+  It should also be noted that shape variation is not equal along both primary dimensions.
+For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first primary dimension.
+Slight differences are seen along the second primary dimension, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first primary dimension.
 
@@
 
-```{r contribution_plots}
+```{r}
 # let's look at all the PCs and how much each original # variable contributes to them
@@
 #view how much variation is described by each PC (ordered from PC1 to last PC)
-# enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
+enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
 
 #get meaningful PCs
-# getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
+getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
 
@@
 # take a look at the plot
-# PC1_contrib
+PC1_contrib
 
@@
 # take a look at the plot
-# PC2_contrib
+PC2_contrib
 
@@
 # examine landmarks with major contribution (%) to PC1 and PC2
-# important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
-# important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
-# important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
+important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
+important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
+important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
 # here is a table that summarises the most important landmarks in the PCA
-# important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
-
-library(cowplot)
-
-# This is a temporary fix just to get the labels looking nice for now; there is most certainly a cleaner way to do this
-# My fix did not work. I will have to create a temp table showing the values for now
-
-# PC1_contrib[["plot_env"]][["xx"]] <- as.factor(c("X Value: Top Left",
-#                                              "Y Value: Top Left",
-#                                               "Y Value: Mid-Upper Left",
-#                                               "X Value: Mid-Lower Left",
-#                                               "Y Value: Mid-Lower Left",
-#                                               "X Value: Bottom Left",
-#                                               "Y Value: Bottom Left",
-#                                               "X Value: Distal Tip",
-#                                               "Y Value: Distal Tip",
-#                                               "X Value: Bottom Right",
-#                                               "Y Value: Bottom Right",
-#                                               "X Value: Mid-Lower Right",
-#                                               "Y Value: Mid-Lower Right",
-#                                               "X Value: Mid-Upper Right",
-#                                               "Y Value: Mid-Upper Right",
-#                                               "X Value: Top Right",
-#                                               "Y Value: Top Right"))
-#
-
-#iris$Species[iris$Species == 'virginica'] <- 'setosa'
-
-#PC1_contrib[["plot_env"]][["data"]][["name"]][PC1_contrib[["plot_env"]][["data"]][["name"]] == '5.X'] <- "X Value: Distal Tip"
-
-#PC1_contrib[["plot_env"]][["data_sum"]][["name"]] <- as.factor(c(1:18))
-
-output_meanings <-tibble(data.frame(c("X Value: Top Left",
-         "Y Value: Top Left",
-         "X Value: Mid-Upper Left",
-         "Y Value: Mid-Upper Left",
-         "X Value: Mid-Lower Left",
-         "Y Value: Mid-Lower Left",
-         "X Value: Bottom Left",
-         "Y Value: Bottom Left",
-         "X Value: Distal Tip",
-         "Y Value: Distal Tip",
-         "X Value: Bottom Right",
-         "Y Value: Bottom Right",
-         "X Value: Mid-Lower Right",
-         "Y Value: Mid-Lower Right",
-         "X Value: Mid-Upper Right",
-         "Y Value: Mid-Upper Right",
-         "X Value: Top Right",
-         "Y Value: Top Right"),
-       c("1X",
-         "1Y",
-         "2X",
-         "2Y",
-         "3X",
-         "3Y",
-         "4X",
-         "4Y",
-         "5X",
-         "5Y",
-         "6X",
-         "6Y",
-         "7X",
-         "7Y",
-         "8X",
-         "8Y",
-         "9X",
-         "9Y")))
-
-output_meanings <- setNames(output_meanings, c("Landmark Value", "Computer Output"))
-
-plot_grid(PC1_contrib, 
-          PC2_contrib)
- output_meanings
+important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
 
 ```
-
-  <!-- Could not figure out how to change the labels on the x axis for the two bar charts, so I included a small table with the human-meaningful values for now-->
 
@@
 ## References
-Archer, W., Djakovic, I., Brenet, M., Bourguignon, L., Presnyakova, D., Schlager, S., Soressi, M., & McPherron, S.
-P.
-(2021).
-Quantifying differences in hominin flaking technologies with 3D shape analysis.
-Journal of Human Evolution, 150, 102912.
-https://doi.org/10.1016/j.jhevol.2020.102912
 
@@
 
-Hoggard, C.
-S., McNabb, J., & Cole, J.
-N.
+Jung, H., & Woo, E.
+J.
 (2019).
-The Application of Elliptic Fourier Analysis in Understanding Biface Shape and Symmetry Through the British Acheulean.
-Journal of Paleolithic Archaeology, 2(2), 115–133.
-https://doi.org/10.1007/s41982-019-00024-6
+Changes in mandibular ramus shape from the Neolithic to modern periods in Korea.
+International Journal of Osteoarchaeology, 29(4), 634–643.
+https://doi.org/10.1002/oa.2759
 
@@
 
-Okumura, M., & Araujo, A.
-G.
-M.
-(2014).
-Long-term cultural stability in hunter–gatherers: A case study using traditional and geometric morphometric analysis of lithic stemmed bifacial points from Southern Brazil.
-Journal of Archaeological Science, 45, 59–71.
-https://doi.org/10.1016/j.jas.2014.02.009
-
-Radinović, M., & Kajtez, I.
-(2021).
-Outlining the knapping techniques: Assessment of the shape and regularity of prismatic blades using elliptic Fourier analysis.
-Journal of Archaeological Science: Reports, 38, 103079.
-https://doi.org/10.1016/j.jasrep.2021.103079
-
-Riede, F., & Pedersen, J. B. (2018). Late Glacial Human Dispersals in Northern Europe and Disequilibrium Dynamics. Human Ecology, 46(5), 621–632. https://doi.org/10.1007/s10745-017-9964-8
-
-Song, M., & Zhong, H.
-(2020).
-Efficient weighted univariate clustering maps outstanding dysregulated genomic zones in human cancers.
-Bioinformatics, 36(20), 5027–5036.
-https://doi.org/10.1093/bioinformatics/btaa613
+Papageorgiou, I. (2020). Ceramic investigation: How to perform statistical analyses. Archaeological and Anthropological Sciences, 12(9), 210. https://doi.org/10.1007/s12520-020-01142-x
 
@@
 https://doi.org/10.1038/s41596-020-0347-z
-
-Wang, H., & Song, M. (2011). Ckmeans.1d.dp: Optimal k-means Clustering in One Dimension by Dynamic Programming. The R Journal, 3(2), 29–33.
 

### 2022-02-25 analysis/code/comparing-diffs.R

@@ -1,0 +1,41 @@
+# How to see rich diffs of two commits of a single R Markdown document in a local Git repo
+
+# https://github.com/lorenzwalthert/gitsum
+library("gitsum")
+library("tidyverse")
+
+# To browse the commits locally, open an RStudio
+# project that is using Git version control, then ...
+
+# Set the path within the project to the Rmd file
+# in the project that we want to compare versions of
+file_path <- "analysis/paper/paper.Rmd"
+
+# Make a data frame that stores the git commit history,
+# one row per commit
+init_gitsum()
+tbl_out <- parse_log_detailed() %>%
+  select(date, author_name, message, hash)
+
+# View this dataframe to pick the specific commits to compare
+View(tbl_out)
+
+# choose two commits to compare, here I've got the most recent
+# one (the last row of the table), and the one immediately before that
+i <- nrow(tbl_out)
+j <- nrow(tbl_out) - 7
+
+# consrtruct the terminal commants using the SHA of the commits we want to
+# compare, and the file path
+git_show_str_1 <- str_glue('git show ', tbl_out[i, ]$hash, ":", file_path)
+git_show_str_2 <- str_glue('git show ', tbl_out[j, ]$hash, ":", file_path)
+
+# get the contents of the file at each commit and
+# store in a vector
+earlier_commit <- system(git_show_str_2, intern = TRUE)
+later_commit <- system(git_show_str_1, intern = TRUE)
+
+# compute differences and visualise
+diffobj::diffChr(earlier_commit,
+                 later_commit,
+                 mode = "sidebyside")

### 2022-02-25 analysis/paper/paper.Rmd
suppress unwanted messages that appear in the docx, minor formatting edits

--- 2022-02-25 analysis/paper/paper.Rmd
@@
   here::here("analysis/data/derived_data/landmarks.tps"),
-  specID = "ID")
+  specID = "ID",
+  warnmsg = FALSE)
 
@@
 
-sup.geo <- gpagen(lands_no_outliers,  ProcD=F, Proj = T)
+sup.geo <- gpagen(lands_no_outliers,  
+                  ProcD=F, 
+                  Proj = T,
+                  print.progress = FALSE)
 
@@
 
-```{r cortex-biplot}
-# Note, trying to fix NA by removing all rows which lack dorsal cortex values. Issue with the initial set of data with NA values being passed through PCA. In the process of fixing
-
-# Adding extra groups needed for biplots, as well as removing the outliers brought up earlier
+```{r cortex-biplot, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
 
@@
 
-<!-- QH: Later on, the language "the visualization above" will be changed to "Figure X." -->
-<!-- QH: Also, on a secondary note, I had to copy the data file into a different location. That means that there are redundant data files that will need to be cleaned up later -->
-
-<!--“biplot above” replace this figure \@ref(fig:code-block-name)  code-block-name is like this ```{r code-block-name} be sure to give them descriptive, simple names -->
-
 Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
@@
 
-```{r excavation-unit-biplot}
+```{r excavation-unit-biplot, fig.cap="PCA biplot for flake shape and excavation unit."}
 # Excavation unit PCA
@@
 ```
+
 <!-- This portion is lifted largely from the ISKM presentation. Might well reword it later, but it will do for now -->
@@
 
-```{r mass-biplot}
+```{r mass-biplot, fig.cap="PCA biplot for flake shape and mass cluster groups"}
 
@@
     theme(plot.title = element_blank())
-
 ```

### 2022-02-25 analysis/paper/paper.Rmd
x-axis tick labels for PCA contribution plots

@@
 
-```{r contribution_plots}
+```{r contribution-plots, fig.cap="Plots showing the contribution of landmarks to the overall shape variation among flakes."}
 # let's look at all the PCs and how much each original # variable contributes to them
@@
 }
-
-# TODO: decode the x-axis labels on the scree plot so we can describe in
-# terms of actual flake dimension, write some sentences to describe and interpret
 
@@
 
-# This is a temporary fix just to get the labels looking nice for now; there is most certainly a cleaner way to do this
-# My fix did not work. I will have to create a temp table showing the values for now
+plot_grid(
 
-# PC1_contrib[["plot_env"]][["xx"]] <- as.factor(c("X Value: Top Left",
-#                                              "Y Value: Top Left",
-#                                               "Y Value: Mid-Upper Left",
-#                                               "X Value: Mid-Lower Left",
-#                                               "Y Value: Mid-Lower Left",
-#                                               "X Value: Bottom Left",
-#                                               "Y Value: Bottom Left",
-#                                               "X Value: Distal Tip",
-#                                               "Y Value: Distal Tip",
-#                                               "X Value: Bottom Right",
-#                                               "Y Value: Bottom Right",
-#                                               "X Value: Mid-Lower Right",
-#                                               "Y Value: Mid-Lower Right",
-#                                               "X Value: Mid-Upper Right",
-#                                               "Y Value: Mid-Upper Right",
-#                                               "X Value: Top Right",
-#                                               "Y Value: Top Right"))
-#
-
-#iris$Species[iris$Species == 'virginica'] <- 'setosa'
-
-#PC1_contrib[["plot_env"]][["data"]][["name"]][PC1_contrib[["plot_env"]][["data"]][["name"]] == '5.X'] <- "X Value: Distal Tip"
-
-#PC1_contrib[["plot_env"]][["data_sum"]][["name"]] <- as.factor(c(1:18))
-
-output_meanings <-tibble(data.frame(c("X Value: Top Left",
+PC1_contrib +
+  scale_x_discrete(labels = 
+  as.factor(c("X Value: Top Left",
               "Y Value: Top Left",
-         "X Value: Mid-Upper Left",
               "Y Value: Mid-Upper Left",
@@
               "X Value: Top Right",
-         "Y Value: Top Right"),
-       c("1X",
-         "1Y",
-         "2X",
-         "2Y",
-         "3X",
-         "3Y",
-         "4X",
-         "4Y",
-         "5X",
-         "5Y",
-         "6X",
-         "6Y",
-         "7X",
-         "7Y",
-         "8X",
-         "8Y",
-         "9X",
-         "9Y")))
-
-output_meanings <- setNames(output_meanings, c("Landmark Value", "Computer Output"))
-
-plot_grid(PC1_contrib, 
-          PC2_contrib)
- output_meanings
+              "Y Value: Top Right",
+              "xx"))),
+          PC2_contrib + 
+    scale_x_discrete(labels = 
+  as.factor(c("X Value: Top Left",
+              "Y Value: Top Left",
+              "Y Value: Mid-Upper Left",
+              "X Value: Mid-Lower Left",
+              "Y Value: Mid-Lower Left",
+              "X Value: Bottom Left",
+              "Y Value: Bottom Left",
+              "X Value: Distal Tip",
+              "Y Value: Distal Tip",
+              "X Value: Bottom Right",
+              "Y Value: Bottom Right",
+              "X Value: Mid-Lower Right",
+              "Y Value: Mid-Lower Right",
+              "X Value: Mid-Upper Right",
+              "Y Value: Mid-Upper Right",
+              "X Value: Top Right",
+              "Y Value: Top Right",
+              "xx")))
+  )
 
@@
 
-  <!-- Could not figure out how to change the labels on the x axis for the two bar charts, so I included a small table with the human-meaningful values for now. This must be fixed later-->
-
-  Figure \@ref(fig:contribution_plots) shows the relative contribution of each of the original 18 variables put through our PCA.
+Figure \@ref(fig:contribution-plots) shows the relative contribution of each of the original 18 variables put through our PCA.
 Looking at the supplementary table, we can see that the first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points.

### 2022-02-25 analysis/paper/paper.Rmd
edit to methods section

@@
 Flakes with no dorsal cortex were categorized as tertiary flakes, those with up to a dorsal cortex percentage of 50% were considered secondary and those with a dorsal cortex value greater than 50% were considered primary (Bradbury & Carr 1995).
+
 Mass clusters were generated using univariate k-means, with the end result being 6 mass clusters, with mass cluster 1 having the least mass on average and 6 having the greatest mass on average.
-
-Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group in minimized.
+Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group is minimized.
 We used the Ckmeans.1d.dp package in R to perform our univariate k-means analysis (Wang & Song, 2011).
-The Ckmeans.1d.dp has been used for univariate clustering in applications outside of archaeology prior to our work (Song & Zhong, 2020).
+The Ckmeans.1d.dp has been used for univariate clustering in applications outside of archaeology (Song & Zhong, 2020).
 
 We initially recorded conventional linear dimension measurements, such as max dimension length and width at various points along the max dimension length, from all flakes recovered from the excavation.
-Then we converted these linear dimensions into landmarks points to represent the size and shape of each artefact.
+Then we converted these linear dimensions into landmarks points to represent the shape of each artefact.
 The specific linear measurements used to create the 2D landmarks were the top horizontal measurement, the max dimension length, width at a quarter of the maximum length, width at half of the maximum length, and width at three quarters of the maximum length.
@@
 
-The outlines formed by the shape landmarks were then normalized for size, orientation, and rotation.
-We used Generalized Procrustes Analysis to strip the flakes in the sample dataset of size, orientation, and rotation, leaving only shape data.
-Generalized Procrustes Analysis (or GPA) is a statistical tool for taking shapes on a 2D plane and normalizing them so as to only examine the absolute shape of a set of landmarks.
-An example of GPA being applied to stone tool landmarks is Okumura & Araujo (2014).
-
-Finally, the shape landmark values were analysed by Principal Components Analysis, a form of exploratory data analysis.
-Principal Component Analysis takes the initial set of data and transforms the initial set of variables into a smaller set of variables whist still preserving most of the variation seen in the initial data set.
-PCA has been used to investigate 2D (Riede & Pedersen, 2018; Hoggard et al., 2019; Radinović & Kajtez, 2021) and 3D (Archer et al., 2021) outlines of stone artefacts in the past.
-Our method for carrying out the PCA on the shape landmarks was inspired by Theska et al.
-(2020), which also performed 2D landmark data analysis using PCA.
-PCA biplots, which plot out the first two primary components, were created to visualize differences based upon secondary categorical variables, namely flake reduction categories, excavation unit, and mass cluster.
-The first two primary components formulated the individual points on biplot, while the secondary variables (flake reduction categories, excavation unit, and mass clusters) were represented by colored outlines on the biplots
-
-<!-- BM: Need to add citations to give the reader further information on GPA and shape analysis in archaeology.
-Need to add a sentence of definitino of what is GPA and what is PCA.
-Partly to guide the reader to more details, if they want it, and partly to demonstrate our expertise with the method -->
-
-<!-- QH: Added more information on GPA and PCA, as well as an addition citation -->
+In analysing the landmarks we followed a conventional shape analysis workflow of Generalized Procrustes Analysis (GPA) followed by Principal Components Analysis (PCA) (Riede & Pedersen, 2018; Hoggard et al., 2019; Radinović & Kajtez, 2021; Okumura & Araujo 2014; Archer et al., 2021; Theska et al.
+2020).
+We used GPA to normalize landmarks for size, orientation, and rotation, leaving only shape data.
+The normalized landmarks were then analysed by PCA, a form of exploratory data analysis.
+Principal Component Analysis takes the initial set of data and transforms and reduces them into a smaller set of variables, whist still preserving most of the variation.
+To explore the output we made PCA biplots of the first two Principal Components to visualize shape variation based upon secondary categorical variables, namely flake reduction categories, excavation units, and mass clusters.
 
@@
 The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
-Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slighly larger range of shape variation.
+Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slightly larger range of shape variation.
 Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
@@
               "Y Value: Top Left",
+              "X Value: Mid-Upper Left",
               "Y Value: Mid-Upper Left",
@@
               "X Value: Top Right",
-              "Y Value: Top Right",
-              "xx"))),
+              "Y Value: Top Right"))),
           PC2_contrib + 
@@
               "X Value: Top Right",
-              "Y Value: Top Right",
-              "xx")))
+              "Y Value: Top Right")))
   )
-
 ```
@@
 This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time.
-By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:mass-biplot), where there is a high degree of shape variation in the lighest flakes.
+By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:mass-biplot), where there is a high degree of shape variation in the lightest flakes.
 This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter.
@@
 
-Looking at \@ref(fig:contribution_plots),
+Looking at \@ref(fig:contribution-plots),
 

### 2022-03-03 analysis/paper/paper.Rmd
Tried to pull all three of the biplots into one plot_grid, with mixed results. Currently in the process of working things out, might not be able to put all three plots into one plot_grid due to the sheer size and complexity of each plot, which does not seem to play well with cowplot.

@@
 # ggplot2-based plot of the PCA results using a factoextra function
-fviz_pca_ind(pca.geo,
+cortex_biplot <- fviz_pca_ind(pca.geo,
              geom.ind = "point", # show points without labels
@@
 # ggplot2-based plot of the PCA results using a factoextra function
-fviz_pca_ind(pca.geo,
+excavation_unit_biplot <- fviz_pca_ind(pca.geo,
              geom.ind = "point", # show points without labels
@@
 # ggplot2-based plot of the PCA results using a factoextra function
-fviz_pca_ind(pca.geo,
+mass_biplot <- fviz_pca_ind(pca.geo,
              geom.ind = "point", # show points without labels
@@
 These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.
+
+```{r}
+library(cowplot)
+
+plot_grid(
+  cortex_biplot, NULL, NULL,  excavation_unit_biplot,  mass_biplot, NULL,
+  ncol = 2,
+  labels = c("A", NULL, NULL,"B","C",NULL),
+  label_size = 12,
+  align = "v",
+  rel_heights = c(1.1,
+                  1,
+                  1.1,
+                  4,
+                  1,
+                  1),
+  rel_widths = c(2,
+                 2,
+                 2,
+                 4,
+                 2,
+                 1))
+
+#cortex_biplot + theme(plot.margin = unit(c(3,3,3,3), "lines"))
+#excavation_unit_biplot
+#mass_biplot
+
+#plot_grid(
+#  plot_a, NULL, plot_b
+#  , nrow = 1
+#  , rel_widths = c(1, 0.05, 1)
+#  , labels = c("A", "", "B")
+#)
+
+    #      axis = "none",
+   #       scale = 1,
+   #       nrow = 3,
+    #      ncol = 1,
+   #       rel_widths = c(1,1,1),
+   #       rel_heights = c(1,1,1)
+
+```
 

### 2022-03-04 analysis/paper/paper.Rmd
Adding in a schematic diagram for the landmarks
Update to results section, moved the biplots and their descriptive text around

@@
 ## Methods
+
+![This is a test.](./test_graphic.png)
 
@@
 
-```{r cortex-biplot, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
+```{r, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
 
@@
 
-Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
-The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
-Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slightly larger range of shape variation.
-Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
-
-Shape variation is not equal along both principal components.
-For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first principal component.
-Slight differences are seen along the second principal component, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first principal component.
-
-```{r excavation-unit-biplot, fig.cap="PCA biplot for flake shape and excavation unit."}
+```{r, fig.cap="PCA biplot for flake shape and excavation unit."}
 # Excavation unit PCA
@@
 
-Figure \@ref(fig:excavation-unit-biplot) shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from.
-Excavation unit one is near the surface, and unit 13 is near the base.
-The general picture here is that shape variation changes little through the deposit.
-Units with fewer flakes, such as 13, show lower shape variation.
-
-```{r mass-biplot, fig.cap="PCA biplot for flake shape and mass cluster groups"}
+```{r, fig.cap="PCA biplot for flake shape and mass cluster groups"}
 
@@
 
-In figure \@ref(fig:mass-biplot) we see that greater shape variation is found in the lightest mass classes.
-The two lightest mass clusters, one and two, are in purple and blue respectively.
-These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.
-
+<!-- Below are my attempts to create a nice looking plot_grid of the three biplots.
+I was not able to get it to display properly, so I am just printing off each biplot individually all in a row as a temporary measure.
+-->
 ```{r}
@@
 
-plot_grid(
-  cortex_biplot, NULL, NULL,  excavation_unit_biplot,  mass_biplot, NULL,
-  ncol = 2,
-  labels = c("A", NULL, NULL,"B","C",NULL),
-  label_size = 12,
-  align = "v",
-  rel_heights = c(1.1,
-                  1,
-                  1.1,
-                  4,
-                  1,
-                  1),
-  rel_widths = c(2,
-                 2,
-                 2,
-                 4,
-                 2,
-                 1))
+# plot_grid(
+#  cortex_biplot, NULL, NULL,  excavation_unit_biplot,  mass_biplot, NULL,
+#  ncol = 2,
+#  labels = c("A", NULL, NULL,"B","C",NULL),
+#  label_size = 12,
+#  align = "v",
+#  rel_heights = c(1.1,
+#                  1,
+#                  1.1,
+#                  4,
+#                  1,
+#                  1),
+#  rel_widths = c(2,
+#                 2,
+#                 2,
+#                 4,
+#                 2,
+#                 1))
 
@@
 ```
+
+```{r cortex-biplot, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
+cortex_biplot
+```
+
+```{r excavation-unit-biplot, fig.cap="PCA biplot for flake shape and excavation unit."}
+excavation_unit_biplot
+```
+
+```{r mass-biplot, fig.cap="PCA biplot for flake shape and mass cluster groups"}
+mass_biplot
+```
+
+Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
+The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
+Secondary flakes, which have been retouched and thus have less dorsal cortex than the primary flakes, show a slightly larger range of shape variation.
+Lastly, the tertiary flakes at the end of the reduction sequence show the greatest amount of shape variation.
+
+Shape variation is not equal along both principal components.
+For flakes in all three dorsal cortex classes, the largest differences between each inter-nested is along the first principal component.
+Slight differences are seen along the second principal component, with an increasing amount of variation as one goes from primary to secondary to tertiary flakes, but these changes are minor when compared to the larger changes seen in the first principal component.
+
+Figure \@ref(fig:excavation-unit-biplot) shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from.
+Excavation unit one is near the surface, and unit 13 is near the base.
+The general picture here is that shape variation changes little through the deposit.
+Units with fewer flakes, such as 13, show lower shape variation.
+
+In Figure \@ref(fig:mass-biplot) we see that greater shape variation is found in the lightest mass classes.
+The two lightest mass clusters, one and two, are in purple and blue respectively.
+These first two mass classes span the largest area on the biplot, signifying a larger degree of variation overall.
 

### 2022-03-04 analysis/paper/paper.Rmd
Reduced axes text size on contribution plots

@@
 
+<!-- QH: Trying to think of a way to make this figure work with the current figure numbering system we have in bookdown. Adding in a number that starts before 1.1 for now -->
 ## Methods
 
-![This is a test.](./test_graphic.png)
+![Figure 1.0.
+Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake.
+A, B, C, D, and E are the top horizontal measurement, width at a quarter of the maximum length, maximum dimension length, width at a quarter of the maximum length, and width at three quarter of the maximum length respectively.
+Points 1 through 9 are all of the landmarks generated from measurements A through E.
+](./test_graphic.png)
 
@@
 
-plot_grid(
-
-PC1_contrib +
+plot_grid(PC1_contrib +
           scale_x_discrete(labels = 
@@
                           "X Value: Top Right",
-              "Y Value: Top Right"))),
+                          "Y Value: Top Right")))+
+             theme(axis.text = element_text(size = 5)),
           PC2_contrib + 
@@
                            "X Value: Top Right",
-              "Y Value: Top Right")))
+                           "Y Value: Top Right")))+
+             theme(axis.text = element_text(size = 5))
   )
+
 ```
@@
 
-Looking at \@ref(fig:contribution-plots),
+Looking at \@ref(fig:contribution-plots), one can see that
 

### 2022-03-04 analysis/paper/paper.Rmd
I noticed an error with our contribution plots. See discussion section for comment.

@@
 
-Looking at \@ref(fig:contribution-plots), one can see that
+ <!-- QH: As I write this I notice a major issue with the axes labels on the contribution plots.
+Instead of substituting the original labels on a one to one basis with our English labels, our plots are just printing out our list of English labels in the order they are listed in our code.
+This has to be fixed.
+I then need to redo my description of the plots, since it is all wrong and based off of the eroneous labels.
+-->
+
+Looking at \@ref(fig:contribution-plots), the landmarks which contributed the most to each principal component were all located on the left side of our artefacts.
+This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the)
+
+one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components.
+Interestingly, the landmarks which contributed the most to each principal component were all located on the left side of our artefacts..
 

### 2022-03-04 analysis/paper/paper.Rmd
Added more text to the discussion section

@@
 - Approach used in the paper is a response to the lack of traditional typological categories in South East Asia
+-Further context from (McAdams et al, 2020)
 
@@
 
+The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies.
+Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017).
+While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis.
+The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of varitiblity increasing as flakes are struck off of increasingly differed cores.
+This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012).
+Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003).
+Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001).
+A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.
+
 # Conclusion
@@
 
+Holdaway, S., & Douglass, M. (2012). A Twenty-First Century Archaeology of Stone Artifacts. Journal of Archaeological Method and Theory, 19(1), 101–131. https://doi.org/10.1007/s10816-011-9103-6
+
+McAdams, C., Morley, M.
+W., Fu, X., Kandyba, A.
+V., Derevianko, A.
+P., Nguyen, D.
+T., Doi, N.
+G., & Roberts, R.
+G.
+(2020).
+The Pleistocene geoarchaeology and geochronology of Con Moong Cave, North Vietnam: Site formation processes and hominin activity in the humid tropics.
+Geoarchaeology, 35(1), 72–97.
+https://doi.org/10.1002/gea.21758
+
+McCall, G. S. (2012). Ethnoarchaeology and the Organization of Lithic Technology. Journal of Archaeological Research, 20(2), 157–203. https://doi.org/10.1007/s10814-011-9056-z
+
+Mijares, A. S. B. (2001). An Expedient Lithic Technology in Northern Luzon (Philippines). Lithic Technology, 26(2), 138–152. https://doi.org/10.1080/01977261.2001.11720983
+
 Mijares, A. 2008. “The late pleistocene to early holocene foragers of northern Luzon.”		Bulletin of the Indo-Pacific Prehistory Association 28: 99-107.
@@
 Riede, F., & Pedersen, J. B. (2018). Late Glacial Human Dispersals in Northern Europe and Disequilibrium Dynamics. Human Ecology, 46(5), 621–632. https://doi.org/10.1007/s10745-017-9964-8
+
+Sillitoe, P., & Hardy, K. (2003). Living Lithics: Ethnoarchaeology in Highland Papua New Guinea. Antiquity, 77(297), 555–566. https://doi.org/10.1017/S0003598X00092619
 

### 2022-03-04 analysis/paper/paper.Rmd
start table for long labels on contribution plot

@@
 
+long_labels <- 
+  tribble(       ~long_labels,             ~short_labels, 
+                 "X Value: Top Left",         "1.X",
+                 "Y Value: Top Left",         "2.X",
+                 "X Value: Mid-Upper Left",    "...",
+                 "Y Value: Mid-Upper Left",    
+                 "X Value: Mid-Lower Left",    
+                 "Y Value: Mid-Lower Left",    
+                 "X Value: Bottom Left",      
+                 "Y Value: Bottom Left",      
+                 "X Value: Distal Tip",        
+                 "Y Value: Distal Tip",       
+                 "X Value: Bottom Right",     
+                 "Y Value: Bottom Right",     
+                 "X Value: Mid-Lower Right",   
+                 "Y Value: Mid-Lower Right",    
+                 "X Value: Mid-Upper Right",    
+                 "Y Value: Mid-Upper Right",    
+                 "X Value: Top Right",          
+                 "Y Value: Top Right",           
+                 )
+
 library(cowplot)

### 2022-03-05 analysis/paper/paper.Rmd
Spellcheck

--- 2022-03-04 analysis/paper/paper.Rmd
@@
 While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis.
-The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of varitiblity increasing as flakes are struck off of increasingly differed cores.
+The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores.
 This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012).

### 2022-03-05 analysis/paper/paper.Rmd
improvements to plots, many minor edits to the text, remove unused code and outdated comments

--- 2022-03-05 analysis/paper/paper.Rmd
@@
 ---
-title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Norther Vietnam"
+title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Northern Vietnam"
 author:
@@
     bookdown::word_document2:
+      number_sections: false
+      global_numbering: true
       fig_caption: yes
@@
 
-<!-- BM: In this section we need to review previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->
+<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->
 
@@
 
-<!-- QH: Trying to think of a way to make this figure work with the current figure numbering system we have in bookdown. Adding in a number that starts before 1.1 for now -->
 ## Methods
 
-![Figure 1.0.
-Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake.
+```{r, schematic-figure, fig.cap="Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake.
 A, B, C, D, and E are the top horizontal measurement, width at a quarter of the maximum length, maximum dimension length, width at a quarter of the maximum length, and width at three quarter of the maximum length respectively.
-Points 1 through 9 are all of the landmarks generated from measurements A through E.
-](./test_graphic.png)
+Points 1 through 9 are all of the landmarks generated from measurements A through E."}
+# knitr::include_graphics(here::here("analysis/flake-measurements-schematic-figure.png"))
+```
 
@@
 
-## PCA Analysis on Dorsal Cortex Groups
+## Principal Components Analyses exploring flake shape variation
 
@@
 
-# TODO: find a meaningful way to break this up into groups
-# pass these groups onto the PCA for drawing ellipses
-
-# this block of code below shows one way to discretize cortex percentage
-# into three groups. We need to explore other cutoffs, and other grouping
-# variables besides dorsal cortex. This code below is a good template for
-# starting these explorations
 ```
 
-```{r, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
+```{r, mass-dorsal-grouping-and-cluster-analysis}
 
@@
                   794,  103 , 598 ,1110 , 906 ,  59 ,   4)) %>% 
- # filter(!is.na(dorsal_cortex_perc)) %>% 
   mutate(dorsal_cortext_group =
@@
   mutate(mass_group = as.factor(test_result_mass$cluster))
+```
+
+```{r pca-cortex-groups}
+
+# to adjust the legend size for all three plots
+legend_key <- 0.25
+legend_title_size <- 6
+legend_text_size <- 4
 
@@
              mean.point = F, #don’t show average point
-             legend.title = "Dorsal Cortex Class") +  #adjust accordingly)
-             coord_fixed(ratio = 1)+ #adjustable for variation-weighted plots
-  theme(plot.title = element_blank())
+             legend.title = "Dorsal\ncortex\nclass") +  #adjust accordingly)
+             coord_fixed(ratio = 1) + #adjustable for variation-weighted plots
+  theme(plot.title = element_blank(),
+        axis.text = element_text(size = legend_text_size),
+        axis.title = element_text(size = legend_title_size),
+        legend.key.size = unit(legend_key, 'cm'), #change legend key size
+        legend.key.height = unit(legend_key, 'cm'), #change legend key height
+        legend.key.width = unit(legend_key, 'cm'), #change legend key width
+        legend.title = element_text(size = legend_title_size), #change legend title font size
+        legend.text = element_text(size = legend_text_size))
 
@@
 
-```{r, fig.cap="PCA biplot for flake shape and excavation unit."}
+```{r, pca-excavation-unit}
 # Excavation unit PCA
@@
                                        colour = "white")) +
-    theme(plot.title = element_blank())
+    theme(plot.title = element_blank(),
+        axis.text = element_text(size = legend_text_size),
+        axis.title = element_text(size = legend_title_size),
+        legend.key.size = unit(legend_key, 'cm'), #change legend key size
+        legend.key.height = unit(legend_key, 'cm'), #change legend key height
+        legend.key.width = unit(legend_key, 'cm'), #change legend key width
+        legend.title = element_text(size = legend_title_size), #change legend title font size
+        legend.text = element_text(size = legend_text_size)) 
 ```
 
-<!-- This portion is lifted largely from the ISKM presentation. Might well reword it later, but it will do for now -->
-
-```{r, fig.cap="PCA biplot for flake shape and mass cluster groups"}
+```{r, pca-mass-cluster-groups}
 
@@
                                        colour = "white")) +
-    theme(plot.title = element_blank())
+      theme(plot.title = element_blank(),
+        axis.text = element_text(size = legend_text_size),
+        axis.title = element_text(size = legend_title_size),
+        legend.key.size = unit(legend_key, 'cm'), #change legend key size
+        legend.key.height = unit(legend_key, 'cm'), #change legend key height
+        legend.key.width = unit(legend_key, 'cm'), #change legend key width
+        legend.title = element_text(size = legend_title_size), #change legend title font size
+        legend.text = element_text(size = legend_text_size)) 
 ```
 
-<!-- Below are my attempts to create a nice looking plot_grid of the three biplots.
-I was not able to get it to display properly, so I am just printing off each biplot individually all in a row as a temporary measure.
--->
-```{r}
+```{r three-pca-biplots, fig.cap="PCA biplots showing (A) variation in flake shape with dorsal cortex class, (B) depth below surface, and (C) mass cluster", fig.height=6}
 library(cowplot)
 
-# plot_grid(
-#  cortex_biplot, NULL, NULL,  excavation_unit_biplot,  mass_biplot, NULL,
-#  ncol = 2,
-#  labels = c("A", NULL, NULL,"B","C",NULL),
-#  label_size = 12,
-#  align = "v",
-#  rel_heights = c(1.1,
-#                  1,
-#                  1.1,
-#                  4,
-#                  1,
-#                  1),
-#  rel_widths = c(2,
-#                 2,
-#                 2,
-#                 4,
-#                 2,
-#                 1))
-
-#cortex_biplot + theme(plot.margin = unit(c(3,3,3,3), "lines"))
-#excavation_unit_biplot
-#mass_biplot
-
-#plot_grid(
-#  plot_a, NULL, plot_b
-#  , nrow = 1
-#  , rel_widths = c(1, 0.05, 1)
-#  , labels = c("A", "", "B")
-#)
-
-    #      axis = "none",
-   #       scale = 1,
-   #       nrow = 3,
-    #      ncol = 1,
-   #       rel_widths = c(1,1,1),
-   #       rel_heights = c(1,1,1)
+plot_grid(
+ cortex_biplot,            # A
+ excavation_unit_biplot,   # B
+ mass_biplot,              # C
+ ncol = 1,
+ align = "v", 
+ axis = "lr",
+ labels = "AUTO")
 
@@
 
-```{r cortex-biplot, fig.cap="PCA biplot for flake shape and dorsal cortex class."}
-cortex_biplot
-```
-
-```{r excavation-unit-biplot, fig.cap="PCA biplot for flake shape and excavation unit."}
-excavation_unit_biplot
-```
-
-```{r mass-biplot, fig.cap="PCA biplot for flake shape and mass cluster groups"}
-mass_biplot
-```
-
-Figure \@ref(fig:cortex-biplot) shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
+Figure \@ref(fig:three-pca-biplots)A shows flake shape (as defined within the first two principal components) in relation to dorsal cortex class.
 The primary class of dorsal cortex, representing initial, unretouched flakes in the reduction sequence, has the smallest range of shape variation.
@@
 
-Figure \@ref(fig:excavation-unit-biplot) shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from.
+Figure \@ref(fig:three-pca-biplots)B shows the flake shapes projected on the first two principal components, and colored according to the excavation unit that they were recovered from.
 Excavation unit one is near the surface, and unit 13 is near the base.
@@
 
-In Figure \@ref(fig:mass-biplot) we see that greater shape variation is found in the lightest mass classes.
+In Figure \@ref(fig:three-pca-biplots)C we see that greater shape variation is found in the lightest mass classes.
 The two lightest mass clusters, one and two, are in purple and blue respectively.
@@
 
-```{r contribution-plots, fig.cap="Plots showing the contribution of landmarks to the overall shape variation among flakes."}
+```{r contribution-plots, fig.cap="Plots showing the contribution of landmarks to the first two Principal Components (A: PC1, B: PC2) that summarise the overall shape variation among flakes.
+The dashed red horizontal line corresponds to the expected value if the contribution from each landmark was uniform.", fig.height = 2}
+
 # let's look at all the PCs and how much each original # variable contributes to them
 
-#view how much variation is described by each PC (ordered from PC1 to last PC)
-# enframe(100*pca.geo$sdev^2/sum(pca.geo$sdev^2))
-
-#get meaningful PCs
-# getMeaningfulPCs(pca.geo$sdev^2, n = length(lands_no_outliers_names))
-
-#estimate contribution (%) of landmarks on PC1
-PC1_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 1)
-# take a look at the plot
-# PC1_contrib
-
-LM_contrib_PC1 <- data.frame(LM = rep(NA, nrow(PC1_contrib$data)/2),
-                             PC1_contrib = rep(NA, nrow(PC1_contrib$data)/2))
-for (i in 1:(nrow(PC1_contrib$data)/2)) {
-  LM_contrib_PC1$PC1_contrib[i] <- sum(PC1_contrib$data[(i*2-1):(i*2),]$contrib)
-  LM_contrib_PC1$LM[i] <- as.character(i)
-}
+# estimate contribution (%) of landmarks on PC1
+PC1_contrib <- fviz_contrib(pca.geo, 
+                            choice = "var", 
+                            axes = 1) +
+   theme(plot.title = element_blank(),
+        axis.text = element_text(size = legend_text_size),
+        axis.title = element_text(size = legend_title_size),
+        legend.key.size = unit(legend_key, 'cm'), #change legend key size
+        legend.key.height = unit(legend_key, 'cm'), #change legend key height
+        legend.key.width = unit(legend_key, 'cm'), #change legend key width
+        legend.title = element_text(size = legend_title_size), #change legend title font size
+        legend.text = element_text(size = legend_text_size))
 
 # estimate contribution (%) of landmarks on PC2
-PC2_contrib <- fviz_contrib(pca.geo, choice = "var", axes = 2)
-# take a look at the plot
-# PC2_contrib
+PC2_contrib <- fviz_contrib(pca.geo, 
+                            choice = "var", 
+                            axes = 2) +
+  theme(plot.title = element_blank(),
+        axis.text = element_text(size = legend_text_size),
+        axis.title = element_text(size = legend_title_size),
+        legend.key.size = unit(legend_key, 'cm'), #change legend key size
+        legend.key.height = unit(legend_key, 'cm'), #change legend key height
+        legend.key.width = unit(legend_key, 'cm'), #change legend key width
+        legend.title = element_text(size = legend_title_size), #change legend title font size
+        legend.text = element_text(size = legend_text_size))
 
-LM_contrib_PC2 <- data.frame(LM = rep(NA, nrow(PC2_contrib$data)/2),
-                             PC2_contrib = rep(NA, nrow(PC2_contrib$data)/2))
-for (i in 1:(nrow(PC2_contrib$data)/2)) {
-  LM_contrib_PC2$PC2_contrib[i] <- sum(PC2_contrib$data[(i*2-1):(i*2),]$contrib)
-  LM_contrib_PC2$LM[i] <- as.character(i)
-}
-
-# examine landmarks with major contribution (%) to PC1 and PC2
-# important_LMs_PC1 <- LM_contrib_PC1[LM_contrib_PC1$PC1_contrib > 100/(nrow(PC1_contrib$data)/2),]
-# important_LMs_PC1[order(decreasing = T, important_LMs_PC1$PC1_contrib),]
-# important_LMs_PC2 <- LM_contrib_PC2[LM_contrib_PC2$PC2_contrib > 100/(nrow(PC2_contrib$data)/2),]
-# here is a table that summarises the most important landmarks in the PCA
-# important_LMs_PC2[order(decreasing = T, important_LMs_PC2$PC2_contrib),]
+# this table will allow us to update the plot axis tick mark labels with human-readable
+# labels
+long_labels <- 
+  tibble::tribble(       ~long_labels,             ~short_labels, 
+                 "X Value: Top Left",         "1.X",
+                 "Y Value: Top Left",         "2.X",
+                 "X Value: Mid-Upper Left",    "...",
+                 "Y Value: Mid-Upper Left",    "...", 
+                 "X Value: Mid-Lower Left",    "...", 
+                 "Y Value: Mid-Lower Left",    "...", 
+                 "X Value: Bottom Left",       "...",
+                 "Y Value: Bottom Left",      "...", 
+                 "X Value: Distal Tip",       "...",  
+                 "Y Value: Distal Tip",       "...", 
+                 "X Value: Bottom Right",     "...", 
+                 "Y Value: Bottom Right",     "...", 
+                 "X Value: Mid-Lower Right",   "...", 
+                 "Y Value: Mid-Lower Right",    "...", 
+                 "X Value: Mid-Upper Right",    "...", 
+                 "Y Value: Mid-Upper Right",    "...", 
+                 "X Value: Top Right",         "...",  
+                 "Y Value: Top Right",         "..."   
+                 )
 
@@
 plot_grid(PC1_contrib +
-          scale_x_discrete(labels = 
-                           as.factor(c("X Value: Top Left",
-                          "Y Value: Top Left",
-                          "X Value: Mid-Upper Left",
-                          "Y Value: Mid-Upper Left",
-                          "X Value: Mid-Lower Left",
-                          "Y Value: Mid-Lower Left",
-                          "X Value: Bottom Left",
-                          "Y Value: Bottom Left",
-                          "X Value: Distal Tip",
-                          "Y Value: Distal Tip",
-                          "X Value: Bottom Right",
-                          "Y Value: Bottom Right",
-                          "X Value: Mid-Lower Right",
-                          "Y Value: Mid-Lower Right",
-                          "X Value: Mid-Upper Right",
-                          "Y Value: Mid-Upper Right",
-                          "X Value: Top Right",
-                          "Y Value: Top Right")))+
              theme(axis.text = element_text(size = 5)),
           PC2_contrib + 
-          scale_x_discrete(labels = 
-                           as.factor(c("X Value: Top Left",
-                           "Y Value: Top Left",
-                           "X Value: Mid-Upper Left",
-                           "Y Value: Mid-Upper Left",
-                           "X Value: Mid-Lower Left",
-                           "Y Value: Mid-Lower Left",
-                           "X Value: Bottom Left",
-                           "Y Value: Bottom Left",
-                           "X Value: Distal Tip",
-                           "Y Value: Distal Tip",
-                           "X Value: Bottom Right",
-                           "Y Value: Bottom Right",
-                           "X Value: Mid-Lower Right",
-                           "Y Value: Mid-Lower Right",
-                           "X Value: Mid-Upper Right",
-                           "Y Value: Mid-Upper Right",
-                           "X Value: Top Right",
-                           "Y Value: Top Right")))+
-             theme(axis.text = element_text(size = 5))
+             theme(axis.text = element_text(size = 5)),
+          labels = "AUTO"
   )
@@
 Figure \@ref(fig:contribution-plots) shows the relative contribution of each of the original 18 variables put through our PCA.
-Looking at the supplementary table, we can see that the first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points.
-The second primary dimension is composed largely of the y-values for bottom-most and upper-most points.
+The first principal component is primarily composed of the x-value for the distal tip landmark, with smaller contributions from the y-values from the two upper-most and mid-upper points.
+The second principal component is composed largely of the y-values for bottom-most and upper-most points.
+
+## Principal Components Analyses exploring core shape variation
 
 # Discussion
-  <!-- This portion partially consists of material from the ISKM presentation.
-Might want to reword things even more.
-Moreover, this section needs citations, which will bolster the overall content of this section.
--->
 
-  Our data visualizations reveal some noteworthy trends regarding the degree of core reduction.
-Figure \@ref(fig:excavation-unit-biplot) shows very little variation over time, with lots of overlap between groups in the biplot.
+Flake shape variation generally shows limited correspondence with other factors.
+Figure \@ref(fig:excavation-unit-biplot)B shows very little variation over time, with lots of overlap between groups in the biplot.
 Units with fewer flakes, such as 13, show lower shape variation.
 This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time.
-By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:mass-biplot), where there is a high degree of shape variation in the lightest flakes.
+By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:excavation-unit-biplot)C, where there is a high degree of shape variation in the lightest flakes.
 This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter.
-This is further supported by Figure \@ref(fig:cortex-biplot), where tertiary flakes with little dorsal cortex show the highest shape variation.
+This is further supported by Figure \@ref(fig:excavation-unit-biplot)A where tertiary flakes with little dorsal cortex show the highest shape variation.
 This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.
@@
 Looking at \@ref(fig:contribution-plots), the landmarks which contributed the most to each principal component were all located on the left side of our artefacts.
-This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the)
-
-one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components.
+This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the) one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components.
 Interestingly, the landmarks which contributed the most to each principal component were all located on the left side of our artefacts..
@@
 Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017).
-While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis.
-The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores.
+<!-- summarise Borel et al.
+to show that we understand what they did and how they came to their conclusion, this might be better done in the introduction section  --> While our analysis does not rely upon outlines, both outlines and landmarks are methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis.
+The high variability in flake shape, especially in the tertiary flakes, suggests that knappers were not looking to systematically create a small set of specific shapes of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores.
 This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012).
-Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003).
-Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001).
-A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.
+Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003), relative to organic technologies.
+Mijares (2001) suggested stone tools were primarily used to create bamboo tools in South East Asia.
+A reliance upon bamboo tools, with stone flakes being an intermediary to produce bamboo tools, may explain the great variability in the lithic assemblages at Mau A.
 

### 2022-03-11 analysis/paper/paper.Rmd
minor edits and review

@@
 We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam.
-We apply a Principal Components Analysis to the flake outlines to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
+We apply a Principal Components Analysis to the flake landmarks to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
 We find that flake shape varies by reduction stage, primarily through differences in flake length and width.
@@
 
-- Stone tools from South East Asia tend to lack distinctive typological categories (Mijares 2008; Borel et al. 2013; Borel et al. 2017)
-- The purpose of this study was to examine the relationship between flake shape and reduction intensity
-- Approach used in the paper is a response to the lack of traditional typological categories in South East Asia
+<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
+
+The late Pleistocene of Southeast Asia is characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
+2013; Borel et al.
+2017).
+This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult.
+Previous papers have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching (Borel et al.
+2013; Borel et al.
+2017).
+The purpose of our study is to examine the relationship between shape and lithic reduction in the assemblage at Mau A, northern Vietnam.
+We ask how flake shape correlates with reduction stage, artefact size and chronology.
+
 -Further context from (McAdams et al, 2020)
@@
 
-<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, not just in Southeast Asia, but anywhere in the world -->
+<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g.
+the previous papers that are most similar to our project  -->
 
@@
                  "X Value: Top Left",         "1.X",
-                 "Y Value: Top Left",         "2.X",
-                 "X Value: Mid-Upper Left",    "...",
-                 "Y Value: Mid-Upper Left",    "...", 
-                 "X Value: Mid-Lower Left",    "...", 
-                 "Y Value: Mid-Lower Left",    "...", 
-                 "X Value: Bottom Left",       "...",
-                 "Y Value: Bottom Left",      "...", 
-                 "X Value: Distal Tip",       "...",  
-                 "Y Value: Distal Tip",       "...", 
-                 "X Value: Bottom Right",     "...", 
-                 "Y Value: Bottom Right",     "...", 
-                 "X Value: Mid-Lower Right",   "...", 
-                 "Y Value: Mid-Lower Right",    "...", 
-                 "X Value: Mid-Upper Right",    "...", 
-                 "Y Value: Mid-Upper Right",    "...", 
-                 "X Value: Top Right",         "...",  
-                 "Y Value: Top Right",         "..."   
+                 "Y Value: Top Left",         "1.Y",
+                 "X Value: Mid-Upper Left",   "2.X",
+                 "Y Value: Mid-Upper Left",   "2.Y", 
+                 "X Value: Mid-Lower Left",   "3.X", 
+                 "Y Value: Mid-Lower Left",   "3.Y", 
+                 "X Value: Bottom Left",      "4.X",
+                 "Y Value: Bottom Left",      "4.Y", 
+                 "X Value: Distal Tip",       "5.X",  
+                 "Y Value: Distal Tip",       "5.Y", 
+                 "X Value: Bottom Right",     "6.X", 
+                 "Y Value: Bottom Right",     "6.Y", 
+                 "X Value: Mid-Lower Right",   "7.X", 
+                 "Y Value: Mid-Lower Right",    "7.Y", 
+                 "X Value: Mid-Upper Right",    "8.X", 
+                 "Y Value: Mid-Upper Right",    "8.Y", 
+                 "X Value: Top Right",         "9.X",  
+                 "Y Value: Top Right",         "9.Y"   
                  )
@@
 This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.
-
- <!-- QH: As I write this I notice a major issue with the axes labels on the contribution plots.
-Instead of substituting the original labels on a one to one basis with our English labels, our plots are just printing out our list of English labels in the order they are listed in our code.
-This has to be fixed.
-I then need to redo my description of the plots, since it is all wrong and based off of the eroneous labels.
--->
+The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies.
+Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017).
+While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis.
+The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores.
+This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012).
+Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003).
+Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001).
+A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.
 
@@
 The high variability in flake shape, especially in the tertiary flakes, suggests that knappers were not looking to systematically create a small set of specific shapes of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores.
+It is also possible that the knappers were aiming to create a wide range of shapes, or that a great number of flakes were created of certain shapes with considerable variation between flakes.
 This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012).

### 2022-03-17 analysis/paper/paper.Rmd
Added in new schematic figure for flake landmarks
temporary subsitution of svg file for png due to issues with labels not printing correctly with the svg file

@@
 Points 1 through 9 are all of the landmarks generated from measurements A through E."}
-# knitr::include_graphics(here::here("analysis/flake-measurements-schematic-figure.png"))
+knitr::include_graphics(here::here("analysis/figures/flake-measurements-schematic-figure-temp.png"))
 ```
@@
 
-plot_grid(PC1_contrib +
+temp <- plot_grid(PC1_contrib +
              theme(axis.text = element_text(size = 5)),

### 2022-03-18 analysis/paper/paper.Rmd
Edit to schematic figure caption

@@
 
-```{r, schematic-figure, fig.cap="Schematic figure showing both the intial measurements that were used to generate our landmarks and the location of our landmarks on a hypothetical flake.
-A, B, C, D, and E are the top horizontal measurement, width at a quarter of the maximum length, maximum dimension length, width at a quarter of the maximum length, and width at three quarter of the maximum length respectively.
-Points 1 through 9 are all of the landmarks generated from measurements A through E."}
+```{r, schematic-figure, fig.cap="Schematic figure showing the location of our landmarks on a hypothetical flake"}
 knitr::include_graphics(here::here("analysis/figures/flake-measurements-schematic-figure-temp.png"))

### 2022-03-18 analysis/paper/paper.Rmd
Slight edit to code creating the contribution bar plots to change the labels "A" and "B" to First and Second Princiap Components respectively. This commit also includes a rewrite of the portion of the discussion section describing the contribution plots; the previous text was inaccurate and based off of incorrect labels.

@@
 
-temp <- plot_grid(PC1_contrib +
+plot_grid(PC1_contrib +
              theme(axis.text = element_text(size = 5)),
@@
              theme(axis.text = element_text(size = 5)),
-          labels = "AUTO"
-  )
+          labels = c("First Principal Component",
+                     "Second Principal Component"),
+          label_size = 6,
+          hjust = -0.3)
 
@@
 
-Looking at \@ref(fig:contribution-plots), the landmarks which contributed the most to each principal component were all located on the left side of our artefacts.
-This means that linking our initial flake measurements (maximum length, top width etc.) to either of the first two principal components is nearly impossible, as every one of our initial flake measurements (with the) one can see that there is no consistent pattern in which certain initial flake measurements  align closely with either the first or second principal components.
-Interestingly, the landmarks which contributed the most to each principal component were all located on the left side of our artefacts..
+Looking at \@ref(fig:contribution-plots), we can see that certain landmarks contributed to each principal component.
+For the first principal component, the distal tip, the top right, and the top left landmarks are the top three contributors.
+We interpret this as indicating that the first principal component is primarily influenced by artefact length and platform breadth.
+The second principal component primarily consists of the bottom left, bottom right, top left, and top right landmarks; one can interpret this as the second principal component taking after overall width.
 

### 2022-03-23 analysis/paper/paper.Rmd
Update to introduction section

@@
 2017).
+Elliptical Fourier analysis was used to approximate shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al.
+2013; Borel et al.
+2017).
+This allowed the outlines of stone tools to be represented as points on a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
+The use of outlines was intentional, as Borel et al.
+(2017) specifically highlighted the dangers of using raw linear measurements, such as length and width, to describe
+
 The purpose of our study is to examine the relationship between shape and lithic reduction in the assemblage at Mau A, northern Vietnam.

### 2022-03-23 analysis/paper/paper.Rmd
Further updates to introduction

@@
 This allowed the outlines of stone tools to be represented as points on a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
-The use of outlines was intentional, as Borel et al.
-(2017) specifically highlighted the dangers of using raw linear measurements, such as length and width, to describe
+The use of outlines and not raw linear measurements was intentional.
+Borel et al.
+(2017) specifically highlighted the dangers of using raw linear measurements, such as length and width, to describe tool morphology, as raw linear measurements can often fail to cover the full variation in shape often seen in stone tools.
 
@@
 We ask how flake shape correlates with reduction stage, artefact size and chronology.
+Our unique innovation consists of an R function whereby sets of linear measurements are converted into coordinates for shape landmarks.
+As far as we know, such an approach has not been undertaken before in the study of stone tool morphology.
+This study can thus be taken as a demonstration of this new approach.
+Our code can be examined at ()
 

### 2022-03-23 analysis/paper/paper.Rmd
Created a 'Paleoenvironmental and Archaeological Context' section, might combine with other sections later on

@@
 ## Introduction
-
-<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
-
+<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g.
+the previous papers that are most similar to our project  -->
 The late Pleistocene of Southeast Asia is characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
@@
 
+## Paleoenvironmental and Archaeological Context
+<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
+
 -Further context from (McAdams et al, 2020)
@@
 Here is a citation [@Marwick2017]
-
-<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g.
-the previous papers that are most similar to our project  -->
 

### 2022-03-24 analysis/paper/paper.Rmd
Additional edits to introduction

@@
 the previous papers that are most similar to our project  -->
+
 The late Pleistocene of Southeast Asia is characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
@@
 This allowed the outlines of stone tools to be represented as points on a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
-The use of outlines and not raw linear measurements was intentional.
+The use of outlines and not the raw linear measurements commonly employed in archaeology was intentional.
 Borel et al.
@@
 Our unique innovation consists of an R function whereby sets of linear measurements are converted into coordinates for shape landmarks.
-As far as we know, such an approach has not been undertaken before in the study of stone tool morphology.
-This study can thus be taken as a demonstration of this new approach.
+Our approach exhibits a sensitivity to morphological variation not seen with raw linear measurements, and can be regarded a method for more usefully employing existing linear measurement data in the study of stone tool morphometrics.
+As far as we know, such an approach has not been undertaken before in the study of stone tool morphology; this paper is demonstration of this new approach.
 Our code can be examined at ()

### 2022-03-25 analysis/paper/paper.Rmd
Added clarification on how k is chosen with univariate k-means
Update to environmental context

@@
 
+While early hominins reached Mainland Southeast Asia as early as 800 ka ago (Forestier et al., 2022), the
+
+Mainland Southeast Asia 
 -Further context from (McAdams et al, 2020)
@@
 Univariate k-means is a non-hierarchical clustering technique that creates a set number of groups equal to k, with points being assigned to groups in such a manner that the total distances between each point and mean of the group is minimized.
+The algorithm used for univariate k-means calculates the optimum number of groups.
 We used the Ckmeans.1d.dp package in R to perform our univariate k-means analysis (Wang & Song, 2011).
@@
 
+Forestier, H., Zhou, Y., Sophady, H., Li, Y., Codeluppi, D., Auetrakulvit, P., & Zeitoun, V.
+(2022).
+The first lithic industry of mainland Southeast Asia: Evidence of the earliest hominin in a tropical context.
+L’Anthropologie, 126(1), 102996.
+https://doi.org/10.1016/j.anthro.2022.102996
+
 Hoggard, C.

### 2022-03-25 analysis/paper/paper.Rmd
New updates and sources for environmental context

@@
 
-While early hominins reached Mainland Southeast Asia as early as 800 ka ago (Forestier et al., 2022), the
+While early hominins reached Mainland Southeast Asia as early as 800 ka ago (Forestier et al., 2022), the oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018).
 
 Mainland Southeast Asia 
--Further context from (McAdams et al, 2020)
+-Further context from (McAdams et al., 2020)
 
@@
 
+Demeter, F., Shackelford, L., Westaway, K., Barnes, L., Duringer, P., Ponche, J.-L., Dumoncel, J., Sénégas, F., Sayavongkhamdy, T., Zhao, J.-X., Sichanthongtip, P., Patole-Edoumba, E., Dunn, T., Zachwieja, A., Coppens, Y., Willerslev, E., & Bacon, A.-M.
+(2017).
+Early Modern Humans from Tam Pà Ling, Laos: Fossil Review and Perspectives.
+Current Anthropology, 58(S17), S527–S538.
+https://doi.org/10.1086/694192
+
 Forestier, H., Zhou, Y., Sophady, H., Li, Y., Codeluppi, D., Auetrakulvit, P., & Zeitoun, V.
@@
 
+Shackelford, L., Demeter, F., Westaway, K., Duringer, P., Ponche, J.-L., Sayavongkhamdy, T., Zhao, J.-X., Barnes, L., Boyon, M., Sichanthongtip, P., Sénégas, F., Patole-Edoumba, E., Coppens, Y., Dumoncel, J., & Bacon, A.-M.
+(2018).
+Additional evidence for early modern human morphological diversity in Southeast Asia at Tam Pa Ling, Laos.
+Quaternary International, 466, 93–106.
+https://doi.org/10.1016/j.quaint.2016.12.002
+
 Sillitoe, P., & Hardy, K. (2003). Living Lithics: Ethnoarchaeology in Highland Papua New Guinea. Antiquity, 77(297), 555–566. https://doi.org/10.1017/S0003598X00092619

### 2022-03-25 analysis/paper/paper.Rmd
Merged the environmental context section with the broader introduction section and added more text and citations

@@
 the previous papers that are most similar to our project  -->
+<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
+
+While early hominins reached Mainland Southeast Asia as early as 800 ka ago (Forestier et al., 2022), the oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018).
+MIS 2 in Northern Vietnam was characterized by sporadic human occupation of caves, localized differences in climate (with shifting cover of rain forest between the moist highlands and the variable lowlands) and the accumulation of mollusk shell middens (McAdams et al., 2020; McAdams et al., 2022).
+Humans in Mainland Southeast Asia (from now on MSEA) would repeatedly reuse caves, accumulating numerous layers of discarded shells, ash, and lithics over multiple periods of occupation, with intervening periods of abandonment and occupation by faunal species such as bats (McAdams et al., 2020).
 
@@
 
-## Paleoenvironmental and Archaeological Context
-<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
-
-While early hominins reached Mainland Southeast Asia as early as 800 ka ago (Forestier et al., 2022), the oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018).
-
-Mainland Southeast Asia 
--Further context from (McAdams et al., 2020)
-
-Here is a citation [@Marwick2017]
-
+<!-- saving this for later:  Here is a citation [@Marwick2017]  -->
 ## Excavations at Mau A
@@
 
+McAdams, C., Morley, M.
+W., Fu, X., Kandyba, A.
+V., Derevianko, A.
+P., Nguyen, D.
+T., Doi, N.
+G., & Roberts, R.
+G.
+(2022).
+Late Pleistocene shell midden microstratigraphy indicates a complex history of human–environment interactions in the uplands of northern Vietnam.
+Philosophical Transactions of the Royal Society B: Biological Sciences, 377(1849), 20200493.
+https://doi.org/10.1098/rstb.2020.0493
+
 McCall, G. S. (2012). Ethnoarchaeology and the Organization of Lithic Technology. Journal of Archaeological Research, 20(2), 157–203. https://doi.org/10.1007/s10814-011-9056-z

### 2022-03-25 analysis/paper/paper.Rmd
Start of conclusion

@@
 
+This work demonstrates the applicability of our new method in identifying the shape changes that accompany lithic reduction within a paleolithic MSEA assemblage.
+Our method for converting linear measurements into shape landmarks demonstrates a sensitivity to morphological diversity that opens the door for further applications.
+As brought up by Borel et al.
+(2017), raw linear measurements rarely capture the full breadth of variation seen in stone tools.
+Our approach of taking linear measurements and converting them into landmarks exhibits a greater sensitivity to shape variation than the use of linear measurements alone.
+This presents the opportunity for linear measurement data (where the measurements cover the full spectrum that we used) to be more usefully employed in the study of stone tool morphometrics.
+While generated landmarks might not be as sensitive as full outlines, linear measurements of stone tools are common in many archaeological studies.
+As a result, there are many opportunities to apply our method to existing linear measurement data.
+Further applications of our method in the study of MSEA paleolithic assemblages is also appropriate.
+
 # Acknowledgements

### 2022-03-25 analysis/paper/paper.Rmd
Removal of a few excess spaces
editing the text, final WI21 meeting

--- 2022-03-25 analysis/paper/paper.Rmd
@@
 
-The late Pleistocene of Southeast Asia is characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
+<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
+
+Stone artefact manufacture in Mainland Southeast Asia (MSEA) may have began as early as 800 ka ago (Forestier et al., 2022), but many of these claims are contested (Marwick et al.
+2022) and it is not until the Late Pleistocene that an indisputably strong signal of artefact production is evidence.
+The oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018).
+By the Last Glacial Maximum at MIS 2 many areas of the region were occupied by stone artefact using forager groups.
+In Northern Vietnam this time was characterized by sporadic human occupation of caves, localized differences in climate (with shifting cover of rain forest between the moist highlands and the variable lowlands) and the accumulation of mollusk shell middens (McAdams et al., 2020; 2022).
+Throughout the Late Pleistocene humans in MSEA would repeatedly reuse caves, accumulating numerous layers of discarded shells, ash, and lithics over multiple periods of occupation, with intervening periods of abandonment and occupation by faunal species such as bats (McAdams et al., 2020).
+
+Lithic technologies in the wider Southeast Asian region during this time are characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
 2013; Borel et al.
 2017).
-This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult.
-Previous papers have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching (Borel et al.
-2013; Borel et al.
-2017).
-Elliptical Fourier analysis was used to approximate shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al.
-2013; Borel et al.
-2017).
-This allowed the outlines of stone tools to be represented as points on a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
-The use of outlines and not the raw linear measurements commonly employed in archaeology was intentional.
+This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult, relative to regions with more visually distinctive technologies such as northern Europe and parts of Africa and South America.
+In their attempt to ...
 Borel et al.
-(2017) specifically highlighted the dangers of using raw linear measurements, such as length and width, to describe tool morphology, as raw linear measurements can often fail to cover the full variation in shape often seen in stone tools.
+(2013; 2017) have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching.
+They used elliptical Fourier analysis to approximate artefact shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al.
+2013; 2017).
+This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
+Borel et al.
+(2017) choose to use of outlines rather than traditional linear measurements because the claims that often fail to cover the full variation in shape often seen in stone tools.
 
@@
 This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.
+
 The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies.
@@
 
-Looking at \@ref(fig:contribution-plots), we can see that certain landmarks contributed to each principal component.
-For the first principal component, the distal tip, the top right, and the top left landmarks are the top three contributors.
-We interpret this as indicating that the first principal component is primarily influenced by artefact length and platform breadth.
-The second principal component primarily consists of the bottom left, bottom right, top left, and top right landmarks; one can interpret this as the second principal component taking after overall width.
+In Figure \@ref(fig:contribution-plots), we can see that the distal tip, the top right, and the top left landmarks are the top three contributors to each principal component.
+We interpret this as indicating that the first principal component is primarily influenced by artefact length and platform width.
+The second principal component primarily consists of the bottom left, bottom right, top left, and top right landmarks; we interpret this as the second principal component representing surface area of the artefact.
 
@@
 # Conclusion
+
+This work demonstrates the applicability of our new method in identifying the shape changes that accompany lithic reduction within a paleolithic MSEA assemblage.
+Our method for converting linear measurements into shape landmarks demonstrates a sensitivity to morphological diversity that opens the door for further applications.
+Our finding support the conclusions of Borel et al.
+(2017) that linear measurements rarely capture the full breadth of variation seen in stone tools.
+Our approach of taking linear measurements and converting them into landmarks gives more options for examining the sensitivity of shape variation to a variety of factors compared to the use of linear measurements alone.
+While the generated landmarks that we use here might not be as sensitive as full artefact outlines, linear measurements of stone tools are simple to collect with basic equipment.
+As a result, there are many opportunities to apply our method to existing linear measurement data on legacy assemblage.
 
@@
 https://doi.org/10.1002/gea.21758
+
+McAdams, C., Morley, M.
+W., Fu, X., Kandyba, A.
+V., Derevianko, A.
+P., Nguyen, D.
+T., Doi, N.
+G., & Roberts, R.
+G.
+(2022).
+Late Pleistocene shell midden microstratigraphy indicates a complex history of human–environment interactions in the uplands of northern Vietnam.
+Philosophical Transactions of the Royal Society B: Biological Sciences, 377(1849), 20200493.
+https://doi.org/10.1098/rstb.2020.0493
 

### 2023-03-22 analysis/paper/paper.Rmd
added new author

@@
 author:
+  - Pham Thanh Son
+      email: thanhson9119@gmail.com
+      institute: [IofA]
+      correspondence: true
   - Quinn Habedank:
@@
 institute:
+  - IofA: Vien Khao co hoc
   - UofO: University of One Place

### 2023-03-22 analysis/paper/paper.Rmd
Update paper.Rmd

--- 2022-03-25 analysis/paper/paper.Rmd
@@
 author:
+  - Pham Thanh Son
+      email: thanhson9119@gmail.com
+      institute: [IofA]
+      correspondence: true
   - Quinn Habedank:
       email: fl@oneoeurhg.edu
-      institute: [UofO]
+      institute: [UofW]
       correspondence: true
   - Ben Marwick:
-      email: fl@another.edu
-      institute: [UofA]
+      email: bmarwick@uw.edu
+      institute: [UofW]
       correspondence: false
 institute:
-  - UofO: University of One Place
+  - IofA: Vien Khao co hoc
+  - UofW: University of Washington 
   - UofA: University of Another Place

### 2023-03-22 analysis/paper/paper.Rmd
remove affiliation

--- 2022-03-25 analysis/paper/paper.Rmd
@@
 author:
+  - Pham Thanh Son
+      email: thanhson9119@gmail.com
+      institute: [IofA]
+      correspondence: true
   - Quinn Habedank:
       email: fl@oneoeurhg.edu
-      institute: [UofO]
+      institute: [UofW]
       correspondence: true
   - Ben Marwick:
-      email: fl@another.edu
-      institute: [UofA]
+      email: bmarwick@uw.edu
+      institute: [UofW]
       correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - IofA: Vien Khao co hoc
+  - UofW: University of Washington 
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2023-03-23 analysis/paper/paper.Rmd
added new data

@@
 author:
-  - Pham Thanh Son
+  - Pham Thanh Son:
       email: thanhson9119@gmail.com

### 2024-03-12 analysis/paper/paper.qmd
migrate from Rmd to qmd

--- 2023-03-22 analysis/paper/paper.Rmd
@@
 ---
-title: "Late Pleistocene stone artefact technology at Mau A, Yen Bai, Northern Vietnam"
+title: "Title Goes Here"
 author:
-  - Pham Thanh Son
+  - Pham Thanh Son:
+      correspondence: "no"
       email: thanhson9119@gmail.com
-      institute: [IofA]
-      correspondence: true
+      orcid: 0000-0003-1689-0557
+      institute:
+        - IofA
   - Quinn Habedank:
-      email: fl@oneoeurhg.edu
-      institute: [UofW]
-      correspondence: true
+      correspondence: "no"
+      institute: UofW
+  - Seungki Kwak:
+      correspondence: "no"
+      institute: fosg
+      orcid: 0000-0003-1689-0559
   - Ben Marwick:
-      email: bmarwick@uw.edu
-      institute: [UofW]
-      correspondence: false
+      correspondence: "yes"
+      orcid: 0000-0001-7879-4531
+      institute: UofW
 institute:
+  - UofW:
+      name: University of Washington 
+      address: Seattle, WA, USA
   - IofA: Vien Khao co hoc
-  - UofW: University of Washington 
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: 
-    bookdown::word_document2:
-      number_sections: false
-      global_numbering: true
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
+title-block-published: "Last updated"  
+date: now
+date-format: long
+format: 
+  docx:
+    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
+execute:
+  echo: false
+  warning: false
+  message: false
+  comment: "#>"
+  fig-path: "../figures/"
+  fig-dpi: 600
+filters:
+  - ../templates/scholarly-metadata.lua
+  - ../templates/author-info-blocks.lua
+  - ../templates/pagebreak.lua
 bibliography: references.bib
@@
 abstract: |
-  Text of abstract
+  Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret.
+As a result, many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly understood.
+We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam.
+We apply a Principal Components Analysis to the flake landmarks to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
+We find that flake shape varies by reduction stage, primarily through differences in flake length and width.
+Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation.
+These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with traditional approaches.
 keywords: |
@@
 
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
-
-```{r setup, echo = FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = "../figures/",
-  dpi = 300
-)
-```
-
 <!-- The actual document text starts here: -->
 
-# Abstract
-
-Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret.
-As a result, many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly understood.
-We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam.
-We apply a Principal Components Analysis to the flake landmarks to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
-We find that flake shape varies by reduction stage, primarily through differences in flake length and width.
-Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation.
-These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with traditional approaches.
-
-## Introduction
 <!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g.
@@
 
-## Methods
+# Methods
 
-```{r, schematic-figure, fig.cap="Schematic figure showing the location of our landmarks on a hypothetical flake"}
+```{r}
+#| label: schematic-figure
+#| fig-cap: "Schematic figure showing the location of our landmarks on a hypothetical flake"
+
 knitr::include_graphics(here::here("analysis/figures/flake-measurements-schematic-figure-temp.png"))
@@
 
-## Results
+# Results
 
@@
 # import the data
-ma_flakes <- readxl::read_excel(here::here("analysis/data/raw_data/UW Fieldschool Mau A 2015 post_excavation_season_April_2016.xlsx"),
-                                col_names = FALSE)
+ma_flakes <- 
+  readxl::read_excel(here::here("analysis/data/raw_data/UW Fieldschool Mau A 2015 post_excavation_season_April_2016.xlsx"), col_names = FALSE)
 
@@
 
-## References
+# References
 
@@
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
@@
 
-```{r colophon, cache = FALSE}
+```{r}
+#| label: colophon
+#| cache: false
+
 # which R packages and versions?

### 2024-03-13 analysis/paper/paper.qmd
writing the intro

@@
 ---
-title: "Title Goes Here"
+title: "Mau A: A brief but intensively occupied Late Pleistocene open-air site on the Red River, northern Vietnam"
 author:
@@
       correspondence: "no"
-      email: thanhson9119@gmail.com
       orcid: 0000-0003-1689-0557
@@
   - UofW:
-      name: University of Washington 
+      name: Department of Anthropology, University of Washington 
       address: Seattle, WA, USA
-  - IofA: Vien Khao co hoc
+  - IofA: 
+      name: Vien Khao co hoc [Vietnamese Institute of Archaeology]
+      address: Hanoi, Vietnam
 title-block-published: "Last updated"  
@@
 abstract: |
-  Flaked stone artefacts from Southeast Asia typically lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret.
-As a result, many of the cultural dynamics of Pleistocene hunter-gatherers of this region are poorly understood.
-We use 2D shape data to hypothesize a relationship between unretouched flake shape and assemblage reduction intensity at Mau A, an early Holocene archaeological site in northern Vietnam.
-We apply a Principal Components Analysis to the flake landmarks to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
+  We report on a stone artefact assemblage from Mau A, an open air site by the Red River in northern Vietnam.
+Artefacts from Mau A, like many sites in mainland Southeast Asia  lack visually distinctive and strongly patterned forms, which can make them challenging to analyze and interpret.
+As a result, many of the dynamics of Pleistocene hunter-gatherers of this region are poorly understood.
+We explore the use of artefact shape data to investigate relationships between  shape and assemblage reduction intensity at Mau A, a brief but intensively occupied site in northern Vietnam.
+We apply a Principal Components Analysis to  two dimensional landmarks of unretouched flakes to investigate shape variation throughout the reduction sequence (measured by dorsal cortex coverage).
 We find that flake shape varies by reduction stage, primarily through differences in flake length and width.
+We use three dimensional shape analysis of cores to explore their variability.
 Our results suggest that flake shape is sensitive to assemblage reduction intensity, and may give useful comparative insights when other attributes show little variation.
-These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with traditional approaches.
+These results are important for understanding stone artefact assemblages from Southeast Asia which often yield little variation when analysed with long-established approaches.
 keywords: |
@@
 
-<!-- BM: In this section we need to review in very precise detail, previous work on investigating flake shape and identifying what affects flake shape, especially the Borel papers, not just in Southeast Asia, but anywhere in the world, Look in Borel to see what papers they cite, and we should summarise the most relevant of those, e.g.
-the previous papers that are most similar to our project  -->
+While the timing of first stone artefact technologies in mainland Southeast Asia (mSEA) is still uncertain, with claims of over 0.7 Ma [@dereviankoDiscoveryBifacialIndustry2018] contested on geoarchaeological grounds [@marwickTektiteGeoarchaeologyMainland2021], there is general consensus on the charater of Pleistocene lithic technologies as dominated by flaked stone tools which rarely have strongly patterned forms [@mijaresMinoriCaveExpedient2002; @borelHowInterpretInformal2013].
+This lack of widely-recognised typo-technological categories for artefacts has slowed progress in understanding the cultural dynamics, relative to regions with more visually distinctive technologies, such as northern Europe and parts of Africa and South America.
+Much recent work in mSEA has focussed on detailed qualitative documentation of small numbers of cores [e.g.
+@forestierReductionSequencesHoabinhian2022], with the far more numerous flake component often left unexamined.
+Meanwhile, novel approaches to quantitative morphometric analyses of lithic assemblages in other regions has demonstrated potential of these methods to address key questions about past human behaviour [e.g.
+@Okumura_Araujo_2014; @riedeQuantitativeAnalysisFinal2024; @archerGeometricMorphometricRelationship2018].
+In island Southeast Asia, applications of this approach have been led by Borel et al.
+[-@borelHowInterpretInformal2013; -@borelStoneToolForms2017] demonstrating the potential for quantiative analysis of flake shape on assemblages from Song Terus, Indonesia.
 
-<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->
+Borel et al.
+[-@borelHowInterpretInformal2013; -@borelStoneToolForms2017] employed multivariate analysis of flake outlines in relation to other variables, such as traces of usewear and retouch.
+They used elliptical Fourier analysis to quantify artefact shape outlines, and summarised shape differences and similarities with Principal Component Analysis.
+This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables.
+Borel et al.
+use outlines rather than traditional linear measurements because these often fail to cover the full variation in shape often seen in stone tools.
+They found that there was no difference in shape between retouched and unretouched artefacts, and no correlation between a specific overall shape and any specific function indicated by usewear analysis [@borelHowInterpretInformal2013; @borelStoneToolForms2017].
 
-Stone artefact manufacture in Mainland Southeast Asia (MSEA) may have began as early as 800 ka ago (Forestier et al., 2022), but many of these claims are contested (Marwick et al.
-2022) and it is not until the Late Pleistocene that an indisputably strong signal of artefact production is evidence.
-The oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018).
-By the Last Glacial Maximum at MIS 2 many areas of the region were occupied by stone artefact using forager groups.
-In Northern Vietnam this time was characterized by sporadic human occupation of caves, localized differences in climate (with shifting cover of rain forest between the moist highlands and the variable lowlands) and the accumulation of mollusk shell middens (McAdams et al., 2020; 2022).
-Throughout the Late Pleistocene humans in MSEA would repeatedly reuse caves, accumulating numerous layers of discarded shells, ash, and lithics over multiple periods of occupation, with intervening periods of abandonment and occupation by faunal species such as bats (McAdams et al., 2020).
+We build on this pioneering work to explore relationships between shape and lithic reduction in mSEA.
+In some traditions of lithic analysis, reduction sequence analysis can mean reconstructing a detailed narrative of a single artefact or handful or artefacts, based on the sequence of flake removals indicated by flake scars on the artefact [e.g.
+@yinghuaNewTechnologicalAnalysis2021 ].
+There are two important limitations to this approach.
+First is that it is wasteful because the majority of the assemblage (e.g.
+the unretouched flakes) are excluded and the information contained in these pieces is not utlized.
+Ethical fieldwork should prioritize maximizing information gained while minimizing destruction.
+Second is that it historically has shown limited potential to provide data to engage with broader anthropological questions about relationships between technology, mobility and risk in hunter-gatherer groups.
+Tackling these questions has been been highly productive in many parts of the world [@holdawaySurfaceStoneArtifact2020; @goldsteinHuntergathererTechnologicalOrganization2022] To address these limitations, our concept of lithic reduction follows Shott "To improve its practice, TO must treat curation as a continuous variable not categorical state, assimilate the time-averaged quality of assemblages as accumulations, reconcile the synchronic behavior it explains with the diachronic accumulations it analyzes, accommodate history as well as adaptation, and transform itself from narrative device to a body of specified, predictive theory.
+"
 
-Lithic technologies in the wider Southeast Asian region during this time are characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al.
-2013; Borel et al.
-2017).
-This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult, relative to regions with more visually distinctive technologies such as northern Europe and parts of Africa and South America.
-In their attempt to ...
-Borel et al.
-(2013; 2017) have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching.
-They used elliptical Fourier analysis to approximate artefact shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al.
-2013; 2017).
-This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear.
-Borel et al.
-(2017) choose to use of outlines rather than traditional linear measurements because the claims that often fail to cover the full variation in shape often seen in stone tools.
+reduction analysis in mSEA is sometimes Inspired by this work, we explore shape-based methods for characterising flake and core variability in mSEA.
 
@@
 <!-- saving this for later:  Here is a citation [@Marwick2017]  -->
+
 ## Excavations at Mau A
@@
 The chaîne opératoire also involved the shaping of thick ovoid cobbles for the production of choppers or chopping-tools and half-cobbles (longitudinally split) that are shaped into tools with transverse cutting edges.
-Flake scar surfaces and edges are fresh and unweathered, indicating _in situ_ production and limited post-depositional movement.
+Flake scar surfaces and edges are fresh and unweathered, indicating *in situ* production and limited post-depositional movement.
 While previous work has claimed the Mau A assemblage is Son Vian (a precurser to Hoabinhian technology, characterised by unifacial flaking of blocky, cubic cobbles)
@@
 
+```{=html}
 <!-- BM: basic stats on raw materials and change in technology by depth to satisfy readers about the basic details of the assemblage,
@@
 <!-- BM: then QH's more complex analysis on flakes after the basic data -->
+```
+<!-- BM: then previous more complex analysis on cores after QH's analysis of flakes -->
 
-<!-- BM: then previous more complex analysis on cores after QH's analysis of flakes -->
+<!-- Eric's work: https://docs.google.com/presentation/d/1BYQBAb2gunTpRa4Z_DaTjbebYQolxIxL/edit#slide=id.p1 -->
+
+<!-- Quinn's work: https://docs.google.com/presentation/d/1kjQ2BLKRSzCvrn6idHLBcRKfuQwczyA5/edit#slide=id.p6 https://drive.google.com/drive/u/1/folders/1tPZNYUZ0WTwWakB_-voJXRcVcTO5iWVO -->
 
@@
 
-source(here::here("analysis/code/999-functions.R"))
+source(here::here("analysis/paper/code/999-functions.R"))
 

### 2024-03-18 analysis/paper/paper.qmd
add site plan and section drawing

@@
         - IofA
+  - Nguyen Thi Mai Huong:
+      correspondence: "no"
+      orcid: 0000-0003-1689-0557
+      institute:
+        - IofA
   - Quinn Habedank:
+      correspondence: "no"
+      institute: UofW
+  - Isabelle Ami Osuna:
       correspondence: "no"
@@
 In some traditions of lithic analysis, reduction sequence analysis can mean reconstructing a detailed narrative of a single artefact or handful or artefacts, based on the sequence of flake removals indicated by flake scars on the artefact [e.g.
-@yinghuaNewTechnologicalAnalysis2021 ].
+@yinghuaNewTechnologicalAnalysis2021].
 There are two important limitations to this approach.
 First is that it is wasteful because the majority of the assemblage (e.g.
-the unretouched flakes) are excluded and the information contained in these pieces is not utlized.
-Ethical fieldwork should prioritize maximizing information gained while minimizing destruction.
+the unretouched flakes) are excluded and the information contained in these pieces is not utilized.
+Ethical fieldwork should prioritize maximizing information gained while minimizing destruction through excavation.
 Second is that it historically has shown limited potential to provide data to engage with broader anthropological questions about relationships between technology, mobility and risk in hunter-gatherer groups.
-Tackling these questions has been been highly productive in many parts of the world [@holdawaySurfaceStoneArtifact2020; @goldsteinHuntergathererTechnologicalOrganization2022] To address these limitations, our concept of lithic reduction follows Shott "To improve its practice, TO must treat curation as a continuous variable not categorical state, assimilate the time-averaged quality of assemblages as accumulations, reconcile the synchronic behavior it explains with the diachronic accumulations it analyzes, accommodate history as well as adaptation, and transform itself from narrative device to a body of specified, predictive theory.
-"
+Tackling these questions has been been highly productive in many parts of the world [@holdawaySurfaceStoneArtifact2020; @goldsteinHuntergathererTechnologicalOrganization2022] To address these limitations, our concept of lithic reduction follows @shottCostsBenefitsTechnological2018 who argues that lithic analysts should "treat curation as a continuous variable, not categorical state, assimilate the time-averaged quality of assemblages as accumulations, reconcile the synchronic behavior it explains with the diachronic accumulations it analyzes, accommodate history as well as adaptation, and transform itself from narrative device to a body of specified, predictive theory".
+Heeding this call, we investigate artefact shape as a continuous variable and investigate how it varies as cores are increasingly reduced in an assembalge representing a relatively short period.
 
-reduction analysis in mSEA is sometimes Inspired by this work, we explore shape-based methods for characterising flake and core variability in mSEA.
-
-The purpose of our study is to examine the relationship between shape and lithic reduction in the assemblage at Mau A, northern Vietnam.
-We ask how flake shape correlates with reduction stage, artefact size and chronology.
-Our unique innovation consists of an R function whereby sets of linear measurements are converted into coordinates for shape landmarks.
-Our approach exhibits a sensitivity to morphological variation not seen with raw linear measurements, and can be regarded a method for more usefully employing existing linear measurement data in the study of stone tool morphometrics.
-As far as we know, such an approach has not been undertaken before in the study of stone tool morphology; this paper is demonstration of this new approach.
-Our code can be examined at ()
-
-<!-- saving this for later:  Here is a citation [@Marwick2017]  -->
+The goal of our study is to examine the relationship between shape and lithic reduction in the assemblage at Mau A, northern Vietnam.
+We ask how flake shape correlates with artefact size and reduction stage.
+We present a method that uses sets of basic linear measurements of artefacts and converts these into coordinates for shape landmarks.
+Our approach exhibits a sensitivity to morphological variation not seen with traditional linear measurements, and can be regarded a method for more usefully employing existing linear measurement data in the study of stone tool morphometrics.
 
@@
 
-Mau A is an open air archaeological site located on the banks above 5 m above the Red River in Yen Bai Province, Northern Vietnam, at the confluence of a small stream entering the river.
-Excavations were conducted in 2015 by a collaboration including researchers from the University of Washington, the Yen Bai Provincial Museum, the Institute of Archaeology in Hanoi, and the Vietnam National University - Social Sciences and Humanities University.
+Mau A is an open air archaeological site located on the banks 5 m above the Red River in Yen Bai Province, Northern Vietnam, at the confluence of a small stream entering the river [@fig-map].
+Excavations were conducted in 2015 by a collaboration including researchers from the University of Washington, the Yen Bai Provincial Museum, the Institute of Archaeology in Hanoi, and the Vietnam National University - Social Sciences and Humanities.
+We resurveyed the site location in 2022 and found that the excavation area had been built over by houses.
 An area of 2 x 2 m was excavated in 13 units of ten centimeters deep, through dense silty clay deposits to a depth of about 1.3 m below the surface.
-Subtle changes in stratigraphy indicated a slightly sloping deposit with four layers, consistent with reports of excavations in this location in the 1980s.
-A small amount of plastic and modern ceramics were found in the uppermost 0.1 m of deposit.
-The most striking feature identified during our excavations was the third layer, which is distinctive as very dense layer of flaked stone artefacts at about 0.6-1.1 m below the surface.
+Subtle changes in stratigraphy indicated a slightly sloping deposit with four layers [@fig-section], consistent with reports of excavations in this location in the 1980s.
+The four layers were identified by the type and amount of artefacts present in each.
+A small amount of plastic and modern ceramics were found in the uppermost 0.1 m of deposit, we identified this is the first layer.
+Below this was a layer where ceramics and plastics were absent, and a low density of flaked stone artefacts were observed.
+The most striking feature identified during our excavations was the third layer, which was distinctive as very dense layer of flaked stone artefacts at about 0.6-1.1 m below the surface.
 No other features were identified during excavation.
+Below the third layer was sterile deposit to the maximum depth of the excavation at 1.3m.
+The excavation stopped due to time limitations, bedrock was not encountered.
+
+```{r}
+#| label: fig-map
+#| fig-cap: "Site plan of the Mau A excavation.
+The lower panel on the right shows the regional context in Souteast Asia.
+The area of the upper panel is indicated by the red rectangle.
+The upper panel shows the topography and rivers of northern Vietnam, indicating the location of Yen Bai, the town where the site is located, and Hanoi, the nearest large city.
+Topographic map of Vietnam created by Sadalmelik with GMT from publicly released GLOBE data, distributed under a CC0 license."
+
+knitr::include_graphics(here::here("analysis/figures/plan-and-section-drawings/Mau_A_plan.png"))
+
+# Note on how I made this figure: the site plan was taken from field measurements.
+The regional map was from wikipedia, and the topo map also from WP.
+I used Inkscape to combine them and clip them to a good fit
+```
+
+```{r}
+#| label: fig-section
+#| fig-cap: "Section drawing of the Mau A excavation.
+The lower panel shows the south and west walls of the excavation.
+All objects in the section are rocks.
+The locations of the four radiocarbon samples are shown by their sample IDs.
+Upper left: excavation in progress by students from the Vietnam National University and the University of Washington.
+Upper right: photograph of the west wall of the excavation."
+
+knitr::include_graphics(here::here("analysis/figures/plan-and-section-drawings/Mau_A_section.png"))
+
+# Note on how I made this figure: the field drawings were imported into Inkscape, then traced.
+```
 
 This dense lithic layer contained the majority of stone artefacts recovered from the site.
-The chaîne opératoire of Mau A lithic technology is composed of unifacial shaping on long cobbles to produce sumatraloids (sumatralith-like pieces).
-The chaîne opératoire also involved the shaping of thick ovoid cobbles for the production of choppers or chopping-tools and half-cobbles (longitudinally split) that are shaped into tools with transverse cutting edges.
-Flake scar surfaces and edges are fresh and unweathered, indicating *in situ* production and limited post-depositional movement.
-While previous work has claimed the Mau A assemblage is Son Vian (a precurser to Hoabinhian technology, characterised by unifacial flaking of blocky, cubic cobbles)
+In brief, the *chaîne opératoire* of Mau A lithic technology began with the aquisitions of ovoid river cobbles from the banks of the Red River.
+This was followed by unifacial shaping on the long sides of flat ovoid cobbles to produce sumatraloids (sumatralith-like pieces).
+A second sequence, operating in parallel, involved the shaping of thick, blocky ovoid cobbles for the production of choppers or chopping-tools.
+Finally, we observed a third sequence resulting in the production of longitudinally split half-cobbles.
+These split cobbles were further shaped into tools with transverse cutting edges.
+Overall, the flake present scar surfaces and edges that are fresh and unweathered, with no concretions, indicating *in situ* production and limited post-depositional movement.
+Flakes are mostly made from quartzite, unretouched, complete, small, thin, and lacking dorsal cortex.
 
-<!-- BM: need to cite The Oxford Handbook of the Archaeology and Anthropology of Hunter-Gatherers, Ha Van Tan 1997, and the Proceedings of the Prehistoric Society -->
+## Chronological and Environmental Context
 
-Four radiocarbon ages were obtained from isolated charcoal fragments in the deposit.
+Four radiocarbon ages were obtained from isolated charcoal fragments in the deposit, due to the absence of features such as hearths.
 The charcoal was sampled to bracket the upper and lower boundaries of the dense lithic layer.
-<!-- add calibrated age numbers here --> The ages indicate a rapid accumulation process for this layer at the terminal Pleistocene.
+The ages indicate a rapid accumulation process for this layer at the terminal Pleistocene.
+The overlap of all of the highest density ranges for each of the four ages in @fig-radiocarbon-ages further shows that these ages are approximately equivalent.
+Consequently, we treat the lithic assemblage as effectively representing a single event of lithic production and discard and we do not investigate change over time at this site.
 
-<!-- BM: add a paragraph on MH's pollen analysis to indicate local palaeoenvironment, and add her as co-author-->
+The radiocabon ages at Mau A situate human occupation in the Bølling-Allerød interstadial (14,321–12,824 years before present), globally a period of warmer conditions.
+Isotope analysis of a speleothem collected from Hoa Huong cave, central Vietnam, indicates a lengthy dry period spanning the Last Glacial Maximum (c.
+20 ka) followed by an abrupt shift to wetter conditions at onset of the Bølling-Allerød interstadial at \~14 ka and persisting into the Holocene [@pattersonGlacialChangesSea2023].
+
+Analysis of pollen from Mau A show results that are broadly consistent with this global and regional situation of wetter conditions.
+Twenty-three sediment samples were collected from square B1 and processed to extract pollen and spores [@sonLithicTechnologyLate2020].
+We found a low concentration of pollen and spores, with arboreal pollen dominant (56%) represented by Poaceae, Typha sp.; Amaranthaceae; Compositae and Cyperaceae.
+Arboreal pollen (17%) included Pinus sp.; Fagaceae; Magnoliaceae, and fern spores is 27%.
+Below 75cm we did not found any arboreal pollen, but in contrast, non-arboreal taxa were found throughout, e.g.
+Poaceae and Compositae.
+Because of the low pollen concentration, it is difficult to discuss the local environmental context of the site, there is only one signal of wet condition based on the appearance of Typha sp.
+pollen at 5-30 cm and 50-65 cm below the surface (Fig 4.8-4.9).
+Overall the pollen data suggest that this location was subject to relatively wet conditions.
+
+Palaeoshoreline reconstructions indicate that current sea levels were established at around 12 ka [@duongHolocenePaleoshorelineChanges2020].
+At the time Mau A was occupied, sea levels may have been up to 70m lower than present [@funabikiHoloceneDeltaPlain2007].
+<!-- Duong to add some text here about what this means for river levels at Mau A -->
+
+```{r}
+#| label: fig-radiocarbon-ages
+#| fig-cap: "Radiocabon ages for Mau A, showing the probability density distribution of individual calibrated dates as ridgeplots.
+The order of the ages on the y-axis follows the depths that the samples were collected from.
+The red lines below each distribution indicte the highest density age range (94% for all except 57% for A1-2-6).
+Full details of the radiocarbon ages can be found in the research compendium."
+
+library(tidyverse)
+
+c14_ages <- 
+read_csv(here::here("analysis/data/derived_data/mau-a-radiocarbon-ages.csv")) %>% 
+  mutate(c14age = `Radiocarbon age BP`,
+         c14std =  `Radiocarbon age error`,
+         prov = str_remove(paste0(Square, "-", Level), ", CH 1|, CH 2")) %>% 
+  mutate(prov = factor(prov, 
+                          levels = c("B1-2-10",
+                                     "A1-2-10",
+                                     "B1-1-6",
+                                     "A1-2-6")))
+
+library(c14bazAAR)
+ rrtools:::quietly( # supress the progress bar
+c14_ages_clean_df_list <- 
+c14_ages %>% 
+  as.c14_date_list() %>% 
+  c14bazAAR::calibrate(choices = c("calprobdistr", "calrange"))
+)
+
+c14_ages_clean_df_list_calprobdistr <- 
+c14_ages_clean_df_list %>% 
+  unnest("calprobdistr")
+
+c14_ages_clean_df_list_calrange <- 
+ c14_ages_clean_df_list %>% 
+  unnest("calrange") %>% 
+  group_by(prov) %>% 
+  filter(dens == max(dens))
+
+library(ggridges)
+
+fig_c14_calibrated <- 
+ggplot(c14_ages_clean_df_list_calprobdistr) +
+  geom_ridgeline(aes(x = -calage + 1950,
+                     y  = prov,
+                     height = density,
+                     scale = 100)) +
+  geom_linerange(data = c14_ages_clean_df_list_calrange,
+                 aes(xmin = -(from - 1950),
+                     xmax = -(to - 1950),
+                     y = prov),
+                 colour = "red",
+                 linewidth = 1) +
+  scale_x_continuous(name = "Age (cal BC)",
+                     sec.axis = sec_axis(~abs(. - 1950),
+                                         name = "Age (cal BP)")) +
+  theme_minimal(base_size = 10)
+
+ggsave(filename = here::here("analysis/figures/fig-c14-calibrated.png"),
+       plot = fig_c14_calibrated,
+       bg = "white",
+       h = 2, 
+       w = 4)
+
+knitr::include_graphics(here::here("analysis/figures/fig-c14-calibrated.png"))
+```
 

### 2024-03-18 analysis/data/raw_data/extract-c14-ages-from-pdf.R

@@ -1,0 +1,26 @@
+
+# this script reads in the PDF and extracts a data frame of the
+# dating results, ready for further analysis
+
+library(tidyverse)
+library(tabulizer) # remotes::install_github(c("ropensci/tabulizerjars", "ropensci/tabulizer"))
+input_file <- str_glue('{here::here()}/analysis/data/raw_data/Results UW-40.pdf')
+pdf_table <- extract_tables(input_file)[[1]] %>% as_data_frame()
+
+pdf_table_nice <-
+  pdf_table[-c(1:7),] %>%  # drop first seven rows that have no data
+  separate(V2, into = c("Square", "Level"), sep = "  ") %>%
+  separate(V4, into = c("pMC", "pMC error"), sep = " ") %>%
+  separate(V5, into = c("Radiocarbon age BP", "Radiocarbon age error"), sep = " ") %>%
+  rename("Lab code" = "V1") %>%
+  rename("13C per mil" = "V3") %>%
+  filter(`Radiocarbon age BP` != "") %>%
+  mutate_at(vars(`13C per mil`,
+                 pMC,
+                 `pMC error`,
+                 `Radiocarbon age BP`,
+                 `Radiocarbon age error`),
+            as.numeric)
+
+write_csv(pdf_table_nice,
+          here::here("analysis/data/derived_data/mau-a-radiocarbon-ages.csv"))
